### 2023-02-18 manuscript/paper.Rmd
Library and manuscript added.

---
title: |
  |  \Large Eras in baseball: Change point analysis fun title. 
author: | 
  | \large Mena Whalen \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`mwhalen3@luc.edu`](mailto:mwhalen3@luc.edu) \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: |
  | Baseball is some weird and wild shit. \vspace{2mm}
  | *Keywords*: change point analysis, baseball, 
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)

gjm <- function(x, color = "red") {
  sprintf("\\textcolor{%s}{%s}", color, x)}
```

```{r pkgs}
library(tidyverse)
theme_set(theme_minimal())
library(Lahman)
```

\newpage

# Introduction {#sec:intro}
@Groothius2017 is the most closely related work to ours.  They use univariate time series of the means and SD's of several statistics (i.e. SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrane Multiplier (LM) unit root test @LeeandStrazicich2003.  Notably they find structural breaks in slugging percentage in 1921 and 1992.  In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).

@Berry1999bridging they talk about bridging eras.

When did the steroids era start: https://www.espn.com/mlb/topics/_/page/the-steroids-era#:~:text=Unlike%20other%20MLB%20%22eras%2C%22,leaguewide%20PED%20testing%20until%202003.

Traditional wisdom:
1900-1919 dead ball era

From Woltring:
"Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras. A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005) (17). This study runs through the 2011 season and a seventh era will be added and labeled the Post Steroid Era (2006-2011)"

https://www.baseball-reference.com/bullpen/Deadball_Era
1901-1920

Mound was lowered in december 1968.  
https://www.baseball-reference.com/bullpen/Pitcher%27s_mound

https://www.espn.com/mlb/story/_/id/33238595/major-league-baseball-stops-testing-players-steroids-nearly-20-years-report-says

https://www.semanticscholar.org/paper/Was-There-a-Structural-Break-in-Barry-Bonds's-Bat-Nieswiadomy-Strazicich/0a32effa10a3d26ea4675efcca3f13229efe0f7d

# Methods
@ChoFryzlewwicz2014 and @Cho2016

R Pacakge: @hdbinseg

# Results
```{r echo = FALSE}
library(tidyverse)
library(janitor)
library(hdbinseg)

teams <- read_csv("../data/Teams.csv") %>% 
  clean_names()

team_names <- teams %>% 
  group_by(franch_id) %>% 
  count() %>% 
  filter(n > 120) %>% 
  pull(franch_id)

# teams %>% 
#   filter(franch_id %in% team_names) %>% 
#   group_by(franch_id) %>% 
#   summarise(min(year_id)) ## filter after 1901

old_teams_full <- teams %>% 
  filter(franch_id %in% team_names & year_id > 1900) %>% 
  select(year_id, franch_id, w, l, r, ab, h, hr, bb, so)

teams_wide <- old_teams_full %>% 
  pivot_wider(id_cols = franch_id, 
              names_from = year_id, 
              values_from = hr)

## find thresholding value for cpt using bootstrapping
thrs <- dcbs.thr(as.matrix(teams_wide[-1]), interval = c(1, dim(teams_wide[-1])[2]))
### if I lower the threshold more changepoints
hr_cpt_dcbs <- dcbs.alg(as.matrix(teams_wide[-1]), cp.type = 1, thr = thrs-1)
hr_cpt_dcbs$ecp
hr_cpt_sbs <- sbs.alg(as.matrix(teams_wide[-1]), cp.type = 1)         
hr_cpt_sbs$ecp

cpt_years <- as.numeric(names(teams_wide[-1])[hr_cpt_dcbs$ecp])

ggplot(old_teams_full, aes(year_id, hr, color = franch_id)) +
  geom_line() +
  geom_vline(xintercept = cpt_years)

```

# Acknowledgements {-}

We thank Michael Lopez for suggesting we do "something with change point analysis".

# Supplementary Material {-}

All code for reproducing the analyses in this paper is publicly available at https://github.com/menawhalen/baseball_cpt

# References

### 2023-02-18 manuscript/paper.tex
Library and manuscript added.

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{article}
\title{~\Large Eras in baseball: Change point analysis fun title.}
\author{\large Mena Whalen \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:mwhalen3@luc.edu}{\texttt{mwhalen3@luc.edu}}
\vspace{-1mm}\\
\strut \\
\large Gregory J. Matthews \vspace{-1.1mm}\\
\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
\normalsize Loyola University Chicago \vspace{-1mm}\\
\normalsize Chicago, IL 60660 \vspace{-1mm}\\
\normalsize \href{mailto:gmatthews1@luc.edu}{\texttt{gmatthews1@luc.edu}}
\vspace{-1mm}}
\date{}

\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  colorlinks=true,
  linkcolor={cyan},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={cyan},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{setspace} \setstretch{1.15} \usepackage{float} \floatplacement{figure}{t}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\begin{document}
\maketitle
\begin{abstract}
Baseball is some weird and wild shit. \vspace{2mm}\\
\emph{Keywords}: change point analysis, baseball,
\end{abstract}

\newpage

\hypertarget{sec:intro}{%
\section{Introduction}\label{sec:intro}}

\protect\hyperlink{ref-Berry1999bridging}{Berry, Reese, and Larkey}
(\protect\hyperlink{ref-Berry1999bridging}{1999}) they talk about
bridging eras.

When did the steroids era start:
\url{https://www.espn.com/mlb/topics/_/page/the-steroids-era\#}:\textasciitilde:text=Unlike\%20other\%20MLB\%20\%22eras\%2C\%22,leaguewide\%20PED\%20testing\%20until\%202003.

Traditional wisdom: 1900-1919 dead ball era

From Woltring: ``Baseball has endured much change over the course of its
history, and because of constant change, the modern era of baseball has
been segmented into six distinct sub-eras. A common list presented at
Baseball-Reference described the eras as the Dead Ball Era (1901-1919),
the Live Ball Era (1920-1941), the Integration Era (1942-1960), the
Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long
Ball/Steroid Era (1994-2005) (17). This study runs through the 2011
season and a seventh era will be added and labeled the Post Steroid Era
(2006-2011)''

\url{https://www.baseball-reference.com/bullpen/Deadball_Era} 1901-1920

Mound was lowered in december 1968.\\
\url{https://www.baseball-reference.com/bullpen/Pitcher\%27s_mound}

\url{https://www.espn.com/mlb/story/_/id/33238595/major-league-baseball-stops-testing-players-steroids-nearly-20-years-report-says}

\hypertarget{methods}{%
\section{Methods}\label{methods}}

\protect\hyperlink{ref-ChoFryzlewwicz2014}{H. Cho and Fryzlewicz}
(\protect\hyperlink{ref-ChoFryzlewwicz2014}{2014}) and
\protect\hyperlink{ref-Cho2016}{H. Cho}
(\protect\hyperlink{ref-Cho2016}{2016})

R Pacakge: \protect\hyperlink{ref-hdbinseg}{Haeran Cho and Fryzlewicz}
(\protect\hyperlink{ref-hdbinseg}{2018})

\hypertarget{results}{%
\section{Results}\label{results}}

\begin{verbatim}
## [1] 20 46 67 94
\end{verbatim}

\begin{verbatim}
##  b 
## 48
\end{verbatim}

\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-1-1} \end{center}

\hypertarget{acknowledgements}{%
\section*{Acknowledgements}\label{acknowledgements}}
\addcontentsline{toc}{section}{Acknowledgements}

We thank Michael Lopez for suggesting we do ``something with change
point analysis.''

\hypertarget{supplementary-material}{%
\section*{Supplementary Material}\label{supplementary-material}}
\addcontentsline{toc}{section}{Supplementary Material}

All code for reproducing the analyses in this paper is publicly
available at \url{https://github.com/menawhalen/baseball_cpt}

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Berry1999bridging}{}}%
Berry, Scott M., C. Shane Reese, and Patrick D. Larkey. 1999.
{``Bridging Different Eras in Sports.''} \emph{Journal of the American
Statistical Association} 94 (447): 661--76.
\url{https://doi.org/10.1080/01621459.1999.10474163}.

\leavevmode\vadjust pre{\hypertarget{ref-Cho2016}{}}%
Cho, H. 2016. {``Change-Point Detection in Panel Data via Double CUSUM
Statistic.''} \emph{Electronic Journal of Statistics} 10: 2000--2038.

\leavevmode\vadjust pre{\hypertarget{ref-hdbinseg}{}}%
Cho, Haeran, and Piotr Fryzlewicz. 2018. \emph{Hdbinseg: Change-Point
Analysis of High-Dimensional Time Series via Binary Segmentation}.
\url{https://CRAN.R-project.org/package=hdbinseg}.

\leavevmode\vadjust pre{\hypertarget{ref-ChoFryzlewwicz2014}{}}%
Cho, H., and P. Fryzlewicz. 2014. {``Multiple-Change-Point Detection for
High Dimensional Time Series via Sparsified Binary Segmentation.''}
\emph{JRSSB} 77: 475--507.

\end{CSLReferences}

\end{document}

### 2023-02-18 hr_cpt.R

@@ -19,25 +19,36 @@ teams %>%
 
 old_teams_full <- teams %>% 
   filter(franch_id %in% team_names & year_id > 1900) %>% 
-  select(year_id, franch_id, w, l, r, ab, h, hr, bb, so)
+  select(year_id, franch_id,hbp, e, sb, w, l, r, ab, h, hr, bb, so,g, attendance,era) %>% 
+  mutate(hr_per_game = hr/g, 
+         so_per_game = so/g,
+         bb_per_game = bb/g,
+         w_per_game = w/g,
+         att_per_game = attendance/g,
+         e_per_game = e/g, 
+         sb_per_game = sb/g, 
+         h_per_game = h/g, 
+         r_per_game = r/g,
+         hbp_per_game = hbp/g)
 
 
 teams_wide <- old_teams_full %>% 
   pivot_wider(id_cols = franch_id, 
               names_from = year_id, 
-              values_from = hr)
+              values_from = hr_per_game)
 
 ## find thresholding value for cpt using bootstrapping
 thrs <- dcbs.thr(as.matrix(teams_wide[-1]), interval = c(1, dim(teams_wide[-1])[2]))
 ### if I lower the threshold more changepoints
-hr_cpt_dcbs <- dcbs.alg(as.matrix(teams_wide[-1]), cp.type = 1, thr = thrs)
+hr_cpt_dcbs <- dcbs.alg(as.matrix(teams_wide[-1]), cp.type = 1, thr = thrs )
 hr_cpt_dcbs$ecp
-hr_cpt_sbs <- sbs.alg(as.matrix(teams_wide[-1]), cp.type = 1)         
+hr_cpt_sbs <- sbs.alg(as.matrix(teams_wide[-1]), cp.type = 1)
 hr_cpt_sbs$ecp
 
 cpt_years <- as.numeric(names(teams_wide[-1])[hr_cpt_dcbs$ecp])
+cpt_years
 
-ggplot(old_teams_full, aes(year_id, hr, color = franch_id)) +
+ggplot(old_teams_full, aes(year_id, hr_per_game, color = franch_id)) +
   geom_line() +
   geom_vline(xintercept = cpt_years)
 

### 2023-02-20 cpts_team_eras.R
R script of each team's change point analysis across multiple metrics

@@ -1,0 +1,71 @@
+library(tidyverse)
+library(janitor)
+library(hdbinseg)
+
+teams <- read_csv("data/Teams.csv") %>% 
+  clean_names()
+
+team_names <- teams %>% 
+  group_by(franch_id) %>% 
+  count() %>% 
+  filter(n > 120) %>% 
+  pull(franch_id)
+
+old_teams_full <- teams %>% 
+  filter(franch_id %in% team_names & year_id > 1900) %>% 
+  select(year_id, franch_id, w, l, r, ab, h, hr, bb, g) %>% 
+  pivot_longer(cols = -c(year_id, franch_id, g), names_to = "variable", values_to = "value") %>% 
+  mutate(game_scale = value/g) %>% 
+  pivot_wider(id_cols = c(year_id, franch_id), names_from = variable, values_from = game_scale)
+
+franchise_names <- unique(old_teams_full$franch_id)
+
+# old_teams_full %>% 
+#   pivot_longer(cols = 3:10, names_to = "stat", values_to = "value") %>% 
+#   filter(franch_id == "NYY") %>% 
+#   select(-franch_id) %>% 
+#   pivot_wider(id_cols = stat, 
+#               names_from = year_id, 
+#               values_from = value)
+
+wide_teams <- lapply(franchise_names, function(i){
+  wide <- old_teams_full %>% 
+    pivot_longer(cols = 3:9, names_to = "stat", values_to = "value") %>% 
+    filter(franch_id == i) %>% 
+    select(-franch_id) %>% 
+    pivot_wider(id_cols = stat, 
+                names_from = year_id, 
+                values_from = value)
+  return(wide)
+})
+
+names(wide_teams) <- franchise_names
+
+## checking about missingness
+lapply(lapply(seq_along(wide_teams), function(x){
+  apply(wide_teams[[x]], MARGIN = 2, function(i) sum(is.na(i)))
+}), sum)  
+## 8 of the teams are missing two years of data from 1911 and 1912 always so
+
+changes_team <- lapply(wide_teams, 
+                  function(x){
+                    ## find thresholding value for cpt using bootstrapping
+                    thrs <- dcbs.thr(as.matrix(x[-1]), interval = c(1, dim(x[-1])[2]))
+                    ### if I lower the threshold more changepoints
+                    hr_cpt_dcbs <- dcbs.alg(as.matrix(x[-1]), cp.type = 1, thr = thrs -3)
[…]

### 2023-02-20 cpts_team_eras.R
changed thresholding parameter to be their choice of bootstrapped threshold

@@ -52,7 +52,7 @@ changes_team <- lapply(wide_teams,
                     ## find thresholding value for cpt using bootstrapping
                     thrs <- dcbs.thr(as.matrix(x[-1]), interval = c(1, dim(x[-1])[2]))
                     ### if I lower the threshold more changepoints
-                    hr_cpt_dcbs <- dcbs.alg(as.matrix(x[-1]), cp.type = 1, thr = thrs -3)
+                    hr_cpt_dcbs <- dcbs.alg(as.matrix(x[-1]), cp.type = 1, thr = thrs)
                     cpt_years <- as.numeric(names(x[-1])[hr_cpt_dcbs$ecp])
                     return(tibble(threshold = thrs,
                                   years = cpt_years))
@@ -68,4 +68,5 @@ old_teams_full %>%
     geom_line() +
     geom_vline(data = teams_cpt_time, aes(xintercept = years)) +
     facet_wrap(~ franch_id, scales = "free_y")
-  
+
+#https://medium.com/@riley.poole/baseball-heroes-of-world-war-ii-9d07e0093b3f  

### 2023-02-21 manuscript/paper.Rmd
Some writing.

@@
 # Introduction {#sec:intro}
+The first professional baseball team in the United States, the Cinicnnati Red Stockings, was formed in 1869.
+Many leagues came and went in the late 1800s, but National League (NL), formed in 1876, emerged as the predominant league of the time.
+A few decades later, the American League (AL) began growing in popularity and eventually reached an agreement with the NL to be the two major leagues of baseball with the winner of each league playing in the World Series starting in 1903.
+
+Throughout the history of baseball, the game has gone through many changes and eras.
+For example, the time period between approximately 1901-1919 is often referred to as the "Dead Ball Era" and was characterized by low scoring games and dominant pitching.
+Another more recent example would be the "Steroid Era" which lasted from approximately 1994 through 2005.
+
+SOME MORE STUFF
+
+The goal of this work is to use multivariate change point analysis to identify change points in MLB as a whole to objectively identify eras in the games history, but also to look for change points to identify team specific eras.
+
+The work that is the most closely related to our own is that of @Groothius2017, which looked for structural breaks (Note: the terms "structural breaks" and "change points" are used interchangeably in this manuscript) in univariate times series.
+
+univariate time series of the means and SD's of several statistics (i.e.
+SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrane Multiplier (LM) unit root test @LeeandStrazicich2003.
+Notably they find structural breaks in slugging percentage in 1921 and 1992.
+In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).
+
+From Woltring:
+"Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras.
+A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005) (17).
+This study runs through the 2011 season and a seventh era will be added and labeled the Post Steroid Era (2006-2011)"
+
+http://mlb.mlb.com/mlb/history/mlb_history_people.jsp?story=com
+
 @Groothius2017 is the most closely related work to ours.
@@
 
+## Results
+
+```{r}
+load("/Users/gregorymatthews/Dropbox/baseball_cpt/data/teams_cpt_time_df.RData")
+teams_cpt_time_df
+
+```
+
+NYY 1918: Babe Ruth got to the Yankees in 1920.  And starting in 1919 they had exactly one losing seasons between 1919 and 1965.  
+BOS 1917, 1929, 1948, 1994: 
+Their last world series win in the 1900s was in 1918.  
+I don't know 1929. 
+1948 there was a big jump in runs?  
+1994: Strike year.
+
+LAD: 1919, 1955
+
+ATL 1918 1949
+
+CHW: 1919: Blaack Sox Scandal
+
+CHC: 1919 1940
+1940: War.
+
+Cin: 1929 1952
+
+CLE 1920 1939
+Cleveland won the world series in 1920. Major change in offensive output.
+
+DET 1927 1940
+
+BAL 1919 1942
+
+SFG 1919 1937
+
+OAK 1920 1936 1951
+
+PHI 1918 1937
+
+PIT 1920 1946
+
+STL 1919 1941
+
+MIN 1919 1942
+
+Pearl Harbor was 1941.  So US was in war in 1942.
+
 # Acknowledgements {-}

### 2023-03-09 manuscript/paper.Rmd
Added some papers.  And lit review

@@
 # Introduction {#sec:intro}
-The first professional baseball team in the United States, the Cinicnnati Red Stockings, was formed in 1869.
+The first professional baseball team in the United States, the Cinicnnati Red Stockings, was formed in 1869 (@BBHOF1869).
 Many leagues came and went in the late 1800s, but National League (NL), formed in 1876, emerged as the predominant league of the time.
@@
 
-Throughout the history of baseball, the game has gone through many changes and eras.
-For example, the time period between approximately 1901-1919 is often referred to as the "Dead Ball Era" and was characterized by low scoring games and dominant pitching.
-Another more recent example would be the "Steroid Era" which lasted from approximately 1994 through 2005.
+Throughout the history of baseball in the United States, the game has gone through many changes and distinct eras.
+For example, the time period between approximately 1900-1919 is often referred to as the "Dead Ball Era" and was marked by low scoring games and dominant pitching.
+Another more recent example would be the "Steroid Era" which lasted from approximately 1994 through 2005 and was characterized by a rapid increase in power hitting largely attributed to players using performance enhancing drugs.
+
+While many baseball writers have attempted to define the different eras of baseball, there has also been some academic work that has sought to empirically define eras in baseball.
+@Groothius2017 looked for structural breaks in univariate times series of performance measures over the period from 1871-2020.
+They analyzed four statistics: slugging percentage (SLG, home run (HR) rate, batting average (BA), and runs batted in (RBI) rate.
+For each of these statistics, the computed the mean and standard deviation (SD) across all players who had at least 100 at bats in a given season to yield a univariate time series for each of these statistics.
+They then used the Lagrange Multiplier (LM) unit root test proposed in @LeeandStrazicich2003 to find structural breaks.
+They identified structural breaks in slugging percentage in 1921 and 1992, the first of which marks the end of the Dead Ball Era and the latter corresponding to the start of the steroid era.
+
+@LeeFort2005 looks for structural changes in competitive balance of the two league American and National.
+Use methods from Andrews1993, Bai1997, 1999 and Bai and Perron 1998 and 2003 to look for break points between 1901 -1999.
+Theuy measure competitive balance in two ways: 1) Noll 1988 and Scully 1989 and 2) Lee (2004).
+They find break point sin competitive balance in 1912, 1926, and 1933 for the NL and in 1926 and 1957 in the AL.
+
+Baseball is not the only sport where this type of analysis has been applied.
+@PalaciosHuerta2004 looked for structural changes in soccer using data from British soccer leagues through 1996.
+@FortLee2007 looked for structural breaks in major North American sports other than baseballl; that is basketball (NBA), hockey (NHL), and American football (NFL).
+WHAT DID THEY FIND?
+
+Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258
+@PalaciosHuerta2004 Structural Breaks in soccer
+
+@Nieswiadomy2012 Was there a structural break in Barry Bonds’s Bat?
+Looking for structural breaks in Bonds OPS times series.
+Use the uni root test on monthly data from 86-07.
+The identify break points in June 1993 and Sept 2000.
+
+In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).
+
+https://onlinelibrary.wiley.com/doi/epdf/10.1111/j.1465-7295.2007.00026.x
+Fort and Lee 2007: Looks for structural breaks in NBA, NHL, and NFL
+
+https://www.baseball-reference.com/bullpen/History_of_baseball_in_the_United_States
+
+https://thesportjournal.org/article/examining-perceptions-of-baseballs-eras/
 
 SOME MORE STUFF
+
+https://pubmed.ncbi.nlm.nih.gov/30747582/?
 
@@
 univariate time series of the means and SD's of several statistics (i.e.
-SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrane Multiplier (LM) unit root test @LeeandStrazicich2003.
+SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrange Multiplier (LM) unit root test @LeeandStrazicich2003.
 Notably they find structural breaks in slugging percentage in 1921 and 1992.
@@
 
+@Woltring2018
 From Woltring:
@@
 
+Deadball era: https://sabr.org/journal/article/the-rise-and-fall-of-the-deadball-era/
+
 http://mlb.mlb.com/mlb/history/mlb_history_people.jsp?story=com
+
+https://www.nytimes.com/2011/09/25/sports/baseball/scoring-in-baseball-returns-to-dead-ball-levels.html?_r=1&
+
+https://bleacherreport.com/articles/1676509-the-evolution-of-the-baseball-from-the-dead-ball-era-through-today
+
+@FortLee2014: 
+Fort R., Lee Y. H. (2006). Stationarity and major league baseball attendance analysis. Journal of Sports Economics, 7, 408–415.
+
+Fort R., Lee Y. H. (2007). Structural change, competitive balance, and the rest of the major leagues. Economic Inquiry, 45, 519–532.
+
+Lee Y. H., Fort R. (2005). Structural change in baseball’s competitive balance: The great depression, team location, and racial integration. Economic Inquiry, 43, 158–169.
+
+@Nieswiadomy2012 Was there a structural break in Barry Bonds’s Bat?
+Looking for structural breaks in Bonds OPS times series.
+Use the uni root test on monthly data from 86-07.
+The identify break points in June 1993 and Sept 2000.
+
+degruyter.com/document/doi/10.1515/1559-0410.1305/html
+
+Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258
+@PalaciosHuerta2004 Structural Breaks in soccer
 
@@
 In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).
+
+Fort, Rodney and Young Hoon Lee (2006). “Stationarity and Major League Baseball Attendance Analysis.” Journal of Sports Economics, 7 (4), 408–415.
+
+Fort, Rodney and Young Hoon Lee (2007). “Structural Change, Competitive Balance, and the Rest of the Major Leagues.” Economic Inquiry, 45 (3), 519-532.
+
+Lee, Young Hoon, and Rodney Fort (2005). “Structural Change in Baseball’s Competitive Balance: The Great Depression, Team Location, and Racial Integration.” Economic Inquiry, 43, 158-169.
+
+Sommers, P. (2008). “The Changing Hitting Performance Profile in Major League Baseball, 1966-2006.” Journal of Sports Economics, 9, 435-440.
 

### 2023-03-09 manuscript/paper.Rmd
Added some more intro stuff.

@@
 
+All of the previous work mentioned here focuses on change point analysis in ***univariate*** time series.
+However, recent methodolgical developments in change point analysis allow for change point analysis in ***multivariate*** time series, which is the focus of this current manuscript.
+This work leverages techniques such as the Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014) to look for change points in Major League Baseball at two main levels.
+First, we look for structural changes in the league as a whole across teams to empirically define different eras in baseball.
+Second, we look for change points within a team to determine eras within a team.
+In essence, this second goal can be used to identify the beginning and end of so called "dynasties", periods of sustained excellent performance by a team.
+In addition, we can also identify the opposite, sustained periods of poor performance which we will term ERAS of SUCKITUDE.
+
+Just Notes: 
 Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258

### 2023-03-15 cpts_all_variables.R
update eras results

@@ -2,6 +2,7 @@ library(tidyverse)
 library(janitor)
 library(hdbinseg)
 
+
 teams <- read_csv("data/Teams.csv") %>% 
   clean_names()
 
@@ -13,30 +14,47 @@ team_names <- teams %>%
 
 old_teams_full <- teams %>% 
   filter(franch_id %in% team_names & year_id > 1900) %>% 
-  select(year_id, franch_id, w, l, r, ab, h, hr, bb, so)
+  select(year_id, franch_id,g, r, ab, sb, hr, bb, so, attendance)
 
 names(old_teams_full)[-c(1:2)]
+## 16 missing values from SO
+apply(old_teams_full, 2,function(x) sum(is.na(x)))
 
-cpt <- function(dataset, var){
-  wide <- dataset %>% 
-    pivot_wider(id_cols = franch_id, 
-                names_from = year_id, 
-                values_from = var)
-  
-  ## find thresholding value for cpt using bootstrapping
-  thrs <- dcbs.thr(as.matrix(wide[-1]), interval = c(1, dim(wide[-1])[2]))
-  ### if I lower the threshold more changepoints
-  hr_cpt_dcbs <- dcbs.alg(as.matrix(wide[-1]), cp.type = 1, thr = thrs)
-  cpt_years <- as.numeric(names(wide[-1])[hr_cpt_dcbs$ecp])
-  return(list(threshold = thrs,
-              years = cpt_years),
-              variable = var)
-  
-}
 
-changes <- lapply(names(old_teams_full)[-c(1:2,10)], 
+imp <- mice::mice(old_teams_full[-c(1:2)])
+teams_full_nomiss <- complete(imp)
+apply(teams_full_nomiss, 2,function(x) sum(is.na(x)))
+
+full_teams <- bind_cols(old_teams_full[1:2],teams_full_nomiss) %>% 
+  mutate(hr_per_game = hr/g, 
+         so_per_game = so/g,
+         bb_per_game = bb/g,
+         att_per_game = attendance/g,
+         sb_per_game = sb/g, 
+         r_per_game = r/g) %>% 
+  select(year_id, franch_id, hr_per_game, so_per_game, bb_per_game, att_per_game, sb_per_game, r_per_game)
+
+
+# cpt <- function(dataset, var){
+#   wide <- dataset %>% 
+#     pivot_wider(id_cols = franch_id, 
+#                 names_from = year_id, 
+#                 values_from = var)
+#   
+#   ## find thresholding value for cpt using bootstrapping
[…]

### 2023-03-15 hr_cpt.R

@@ -19,7 +19,7 @@ teams %>%
 
 old_teams_full <- teams %>% 
   filter(franch_id %in% team_names & year_id > 1900) %>% 
-  select(year_id, franch_id,hbp, e, sb, w, l, r, ab, h, hr, bb, so,g, attendance,era) %>% 
+  select(year_id, franch_id,hbp, e, sb, l, r, ab, h, hr, bb, so,g, attendance,era) %>% 
   mutate(hr_per_game = hr/g, 
          so_per_game = so/g,
          bb_per_game = bb/g,
@@ -35,12 +35,12 @@ old_teams_full <- teams %>%
 teams_wide <- old_teams_full %>% 
   pivot_wider(id_cols = franch_id, 
               names_from = year_id, 
-              values_from = hr_per_game)
+              values_from = att_per_game)
 
 ## find thresholding value for cpt using bootstrapping
 thrs <- dcbs.thr(as.matrix(teams_wide[-1]), interval = c(1, dim(teams_wide[-1])[2]))
 ### if I lower the threshold more changepoints
-hr_cpt_dcbs <- dcbs.alg(as.matrix(teams_wide[-1]), cp.type = 1, thr = thrs )
+hr_cpt_dcbs <- dcbs.alg(as.matrix(teams_wide[-1]), cp.type = 1, thr = thrs)
 hr_cpt_dcbs$ecp
 hr_cpt_sbs <- sbs.alg(as.matrix(teams_wide[-1]), cp.type = 1)
 hr_cpt_sbs$ecp
@@ -48,7 +48,9 @@ hr_cpt_sbs$ecp
 cpt_years <- as.numeric(names(teams_wide[-1])[hr_cpt_dcbs$ecp])
 cpt_years
 
-ggplot(old_teams_full, aes(year_id, hr_per_game, color = franch_id)) +
+ggplot(old_teams_full, aes(year_id, att_per_game, color = franch_id)) +
   geom_line() +
   geom_vline(xintercept = cpt_years)
+
+
 

### 2023-03-15 cpts_all_variables.R
adding second order chpt

@@ -62,7 +62,7 @@ changes <- lapply(names(full_teams)[-c(1:2)],
                     ## find thresholding value for cpt using bootstrapping
                     thrs <- dcbs.thr(as.matrix(wide[-1]), interval = c(1, dim(wide[-1])[2]))
                     ### if I lower the threshold more changepoints
-                    cpt_dcbs <- dcbs.alg(as.matrix(wide[-1]), cp.type = 1, thr = thrs-1)
+                    cpt_dcbs <- dcbs.alg(as.matrix(wide[-1]), cp.type = 1, thr = thrs)
                     cpt_years <- as.numeric(names(wide[-1])[cpt_dcbs$ecp])
                     return(tibble(threshold = thrs,
                                 years = cpt_years,
@@ -70,9 +70,28 @@ changes <- lapply(names(full_teams)[-c(1:2)],
                     
                   })
 
+changes_var <- lapply(names(full_teams)[-c(1:2)], 
+                  function(i){
+                    wide <- full_teams %>% 
+                      pivot_wider(id_cols = franch_id, 
+                                  names_from = year_id, 
+                                  values_from = i)
+                    
+                    ## find thresholding value for cpt using bootstrapping
+                    thrs <- dcbs.thr(as.matrix(wide[-1]), interval = c(1, dim(wide[-1])[2]))
+                    ### if I lower the threshold more changepoints
+                    cpt_dcbs <- dcbs.alg(as.matrix(wide[-1]), cp.type = 2, thr = thrs)
+                    cpt_years <- as.numeric(names(wide[-1])[cpt_dcbs$ecp])
+                    return(tibble(threshold = thrs,
+                                  years = cpt_years,
+                                  variable = i))
+                    
+                  })
+
 time_cpts <- changes %>% 
   tibble() %>% 
-  unnest(cols = c(.)) 
+  unnest(cols = c(.)) %>% 
+  mutate(change = "mean")
 
 
 full_teams %>% 
@@ -83,3 +102,17 @@ full_teams %>%
   facet_wrap(~ variable, scales = "free_y")
 
 
+time_cpts_var <- changes_var %>% 
+  tibble() %>% 
+  unnest(cols = c(.)) %>% 
+  mutate(change = "variance")
+
+
+bind_rows(time_cpts, time_cpts_var)
+
+full_teams %>% 
[…]

### 2023-03-15 manuscript/paper.tex

@@
 \usepackage[margin=1in]{geometry}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 
+The first professional baseball team in the United States, the
+Cinicnnati Red Stockings, was formed in 1869
+(\protect\hyperlink{ref-BBHOF1869}{Rothenberg}
+(\protect\hyperlink{ref-BBHOF1869}{n.d.})). Many leagues came and went
+in the late 1800s, but National League (NL), formed in 1876, emerged as
+the predominant league of the time. A few decades later, the American
+League (AL) began growing in popularity and eventually reached an
+agreement with the NL to be the two major leagues of baseball with the
+winner of each league playing in the World Series starting in 1903.
+
+Throughout the history of baseball in the United States, the game has
+gone through many changes and distinct eras. For example, the time
+period between approximately 1900-1919 is often referred to as the
+``Dead Ball Era'' and was marked by low scoring games and dominant
+pitching. Another more recent example would be the ``Steroid Era'' which
+lasted from approximately 1994 through 2005 and was characterized by a
+rapid increase in power hitting largely attributed to players using
+performance enhancing drugs.
+
+While many baseball writers have attempted to define the different eras
+of baseball, there has also been some academic work that has sought to
+empirically define eras in baseball.
+\protect\hyperlink{ref-Groothius2017}{Groothius, Rotthoff, and
+Strazicich} (\protect\hyperlink{ref-Groothius2017}{2017}) looked for
+structural breaks in univariate times series of performance measures
+over the period from 1871-2020. They analyzed four statistics: slugging
+percentage (SLG, home run (HR) rate, batting average (BA), and runs
+batted in (RBI) rate. For each of these statistics, the computed the
+mean and standard deviation (SD) across all players who had at least 100
+at bats in a given season to yield a univariate time series for each of
+these statistics. They then used the Lagrange Multiplier (LM) unit root
+test proposed in
+(\protect\hyperlink{ref-LeeandStrazicich2003}{\textbf{LeeandStrazicich2003?}})
+to find structural breaks. They identified structural breaks in slugging
+percentage in 1921 and 1992, the first of which marks the end of the
+Dead Ball Era and the latter corresponding to the start of the steroid
+era.
+
+\protect\hyperlink{ref-LeeFort2005}{H. and R.}
+(\protect\hyperlink{ref-LeeFort2005}{2005}) looks for structural changes
+in competitive balance of the two league American and National. Use
+methods from Andrews1993, Bai1997, 1999 and Bai and Perron 1998 and 2003
+to look for break points between 1901 -1999. Theuy measure competitive
+balance in two ways: 1) Noll 1988 and Scully 1989 and 2) Lee (2004).
+They find break point sin competitive balance in 1912, 1926, and 1933
+for the NL and in 1926 and 1957 in the AL.
+
+Baseball is not the only sport where this type of analysis has been
+applied. \protect\hyperlink{ref-PalaciosHuerta2004}{I.}
+(\protect\hyperlink{ref-PalaciosHuerta2004}{2004}) looked for structural
+changes in soccer using data from British soccer leagues through 1996.
+\protect\hyperlink{ref-FortLee2007}{Fort R.}
+(\protect\hyperlink{ref-FortLee2007}{2007}) looked for structural breaks
+in major North American sports other than baseballl; that is basketball
+(NBA), hockey (NHL), and American football (NFL). WHAT DID THEY FIND?
+
+All of the previous work mentioned here focuses on
+\textbf{\emph{univariate}} In contrast to their methods, we use more
+modern methods Double CUSUM Binary Segmentation algorithm
+(\protect\hyperlink{ref-Cho2016}{H. Cho}
+(\protect\hyperlink{ref-Cho2016}{2016})) and the Sparsified Binary
+Segmentation algorithm (\protect\hyperlink{ref-ChoFryzlewwicz2014}{H.
+Cho and Fryzlewicz} (\protect\hyperlink{ref-ChoFryzlewwicz2014}{2014})).
+
+Just Notes: Palacios-Huerta I. (2004). Structural changes during a
+Century of the World's most popular sport. Statistical Methods and
+Applications, 13, 241--258
+\protect\hyperlink{ref-PalaciosHuerta2004}{I.}
+(\protect\hyperlink{ref-PalaciosHuerta2004}{2004}) Structural Breaks in
+soccer
+
+\protect\hyperlink{ref-Nieswiadomy2012}{Nieswiadomy}
+(\protect\hyperlink{ref-Nieswiadomy2012}{2012}) Was there a structural
+break in Barry Bonds's Bat? Looking for structural breaks in Bonds OPS
+times series. Use the uni root test on monthly data from 86-07. The
+identify break points in June 1993 and Sept 2000.
+
+In contrast to their methods, we use more modern methods Double CUSUM
+Binary Segmentation algorithm (\protect\hyperlink{ref-Cho2016}{H. Cho}
+(\protect\hyperlink{ref-Cho2016}{2016})) and the Sparsified Binary
+Segmentation algorithm (\protect\hyperlink{ref-ChoFryzlewwicz2014}{H.
+Cho and Fryzlewicz} (\protect\hyperlink{ref-ChoFryzlewwicz2014}{2014})).
+
+\url{https://onlinelibrary.wiley.com/doi/epdf/10.1111/j.1465-7295.2007.00026.x}
+Fort and Lee 2007: Looks for structural breaks in NBA, NHL, and NFL
+
+\url{https://www.baseball-reference.com/bullpen/History_of_baseball_in_the_United_States}
+
+\url{https://thesportjournal.org/article/examining-perceptions-of-baseballs-eras/}
+
+SOME MORE STUFF
+
+\url{https://pubmed.ncbi.nlm.nih.gov/30747582/}?
+
+The goal of this work is to use multivariate change point analysis to
+identify change points in MLB as a whole to objectively identify eras in
+the games history, but also to look for change points to identify team
+specific eras.
+
+The work that is the most closely related to our own is that of
+\protect\hyperlink{ref-Groothius2017}{Groothius, Rotthoff, and
+Strazicich} (\protect\hyperlink{ref-Groothius2017}{2017}), which looked
+for structural breaks (Note: the terms ``structural breaks'' and
+``change points'' are used interchangeably in this manuscript) in
+univariate times series.
+
+univariate time series of the means and SD's of several statistics
+(i.e.~SLUG, HR rate, RBI rate, and batting average) to look for
+structural breaks via the Lagrange Multiplier (LM) unit root test
+(\protect\hyperlink{ref-LeeandStrazicich2003}{\textbf{LeeandStrazicich2003?}}).
+Notably they find structural breaks in slugging percentage in 1921 and
+1992. In contrast to their methods, we use more modern methods Double
+CUSUM Binary Segmentation algorithm (\protect\hyperlink{ref-Cho2016}{H.
+Cho} (\protect\hyperlink{ref-Cho2016}{2016})) and the Sparsified Binary
+Segmentation algorithm (\protect\hyperlink{ref-ChoFryzlewwicz2014}{H.
+Cho and Fryzlewicz} (\protect\hyperlink{ref-ChoFryzlewwicz2014}{2014})).
+
+\protect\hyperlink{ref-Woltring2018}{Woltring and Jubenville}
+(\protect\hyperlink{ref-Woltring2018}{2018}) From Woltring: ``Baseball
+has endured much change over the course of its history, and because of
+constant change, the modern era of baseball has been segmented into six
+distinct sub-eras. A common list presented at Baseball-Reference
+described the eras as the Dead Ball Era (1901-1919), the Live Ball Era
+(1920-1941), the Integration Era (1942-1960), the Expansion Era
+(1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid
+Era (1994-2005) (17). This study runs through the 2011 season and a
+seventh era will be added and labeled the Post Steroid Era (2006-2011)''
+
+Deadball era:
+\url{https://sabr.org/journal/article/the-rise-and-fall-of-the-deadball-era/}
+
+\url{http://mlb.mlb.com/mlb/history/mlb_history_people.jsp?story=com}
+
+\url{https://www.nytimes.com/2011/09/25/sports/baseball/scoring-in-baseball-returns-to-dead-ball-levels.html?_r=1\&}
+
+\url{https://bleacherreport.com/articles/1676509-the-evolution-of-the-baseball-from-the-dead-ball-era-through-today}
+
+(\protect\hyperlink{ref-FortLee2014}{\textbf{FortLee2014?}}): Fort R.,
+Lee Y. H. (2006). Stationarity and major league baseball attendance
+analysis. Journal of Sports Economics, 7, 408--415.
+
+Fort R., Lee Y. H. (2007). Structural change, competitive balance, and
+the rest of the major leagues. Economic Inquiry, 45, 519--532.
+
+Lee Y. H., Fort R. (2005). Structural change in baseball's competitive
+balance: The great depression, team location, and racial integration.
+Economic Inquiry, 43, 158--169.
+
+\protect\hyperlink{ref-Nieswiadomy2012}{Nieswiadomy}
+(\protect\hyperlink{ref-Nieswiadomy2012}{2012}) Was there a structural
+break in Barry Bonds's Bat? Looking for structural breaks in Bonds OPS
+times series. Use the uni root test on monthly data from 86-07. The
+identify break points in June 1993 and Sept 2000.
+
+degruyter.com/document/doi/10.1515/1559-0410.1305/html
+
+Palacios-Huerta I. (2004). Structural changes during a Century of the
+World's most popular sport. Statistical Methods and Applications, 13,
+241--258 \protect\hyperlink{ref-PalaciosHuerta2004}{I.}
+(\protect\hyperlink{ref-PalaciosHuerta2004}{2004}) Structural Breaks in
+soccer
+
+\protect\hyperlink{ref-Groothius2017}{Groothius, Rotthoff, and
+Strazicich} (\protect\hyperlink{ref-Groothius2017}{2017}) is the most
+closely related work to ours. They use univariate time series of the
+means and SD's of several statistics (i.e.~SLUG, HR rate, RBI rate, and
+batting average) to look for structural breaks via the Lagrane
+Multiplier (LM) unit root test
+(\protect\hyperlink{ref-LeeandStrazicich2003}{\textbf{LeeandStrazicich2003?}}).
+Notably they find structural breaks in slugging percentage in 1921 and
+1992. In contrast to their methods, we use more modern methods Double
+CUSUM Binary Segmentation algorithm (\protect\hyperlink{ref-Cho2016}{H.
+Cho} (\protect\hyperlink{ref-Cho2016}{2016})) and the Sparsified Binary
+Segmentation algorithm (\protect\hyperlink{ref-ChoFryzlewwicz2014}{H.
+Cho and Fryzlewicz} (\protect\hyperlink{ref-ChoFryzlewwicz2014}{2014})).
+
+Fort, Rodney and Young Hoon Lee (2006). ``Stationarity and Major League
+Baseball Attendance Analysis.'' Journal of Sports Economics, 7 (4),
+408--415.
+
+Fort, Rodney and Young Hoon Lee (2007). ``Structural Change, Competitive
+Balance, and the Rest of the Major Leagues.'' Economic Inquiry, 45 (3),
+519-532.
+
+Lee, Young Hoon, and Rodney Fort (2005). ``Structural Change in
+Baseball's Competitive Balance: The Great Depression, Team Location, and
+Racial Integration.'' Economic Inquiry, 43, 158-169.
+
+Sommers, P. (2008). ``The Changing Hitting Performance Profile in Major
+League Baseball, 1966-2006.'' Journal of Sports Economics, 9, 435-440.
+
 \protect\hyperlink{ref-Berry1999bridging}{Berry, Reese, and Larkey}
@@
 
+\url{https://www.semanticscholar.org/paper/Was-There-a-Structural-Break-in-Barry-Bonds's-Bat-Nieswiadomy-Strazicich/0a32effa10a3d26ea4675efcca3f13229efe0f7d}
+
 \hypertarget{methods}{%
@@
 \begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-1-1} \end{center}
+
+\hypertarget{results-1}{%
+\subsection{Results}\label{results-1}}
+
+\begin{longtable}[]{@{}lrr@{}}
+\toprule
+franch\_id & threshold & years \\
+\midrule
+\endhead
+NYY & 4.801818 & 1918 \\
+BOS & 4.427752 & 1917 \\
+BOS & 4.427752 & 1929 \\
+BOS & 4.427752 & 1948 \\
+BOS & 4.427752 & 1994 \\
+LAD & 4.776867 & 1919 \\
+LAD & 4.776867 & 1955 \\
+ATL & 5.393778 & 1918 \\
+ATL & 5.393778 & 1949 \\
+CHW & 4.326192 & 1918 \\
+CHC & 4.277577 & 1919 \\
+CHC & 4.277577 & 1940 \\
+CIN & 4.249476 & 1929 \\
+CIN & 4.249476 & 1952 \\
+CLE & 4.405339 & 1920 \\
+CLE & 4.405339 & 1939 \\
+DET & 3.926054 & 1927 \\
+DET & 3.926054 & 1940 \\
+BAL & 4.641706 & 1919 \\
+BAL & 4.641706 & 1942 \\
+SFG & 4.319634 & 1919 \\
+SFG & 4.319634 & 1937 \\
+OAK & 3.819294 & 1920 \\
+OAK & 3.819294 & 1936 \\
+OAK & 3.819294 & 1951 \\
+PHI & 4.123104 & 1918 \\
+PHI & 4.123104 & 1937 \\
+PIT & 4.175793 & 1920 \\
+PIT & 4.175793 & 1946 \\
+STL & 5.444239 & 1919 \\
+STL & 5.444239 & 1941 \\
+MIN & 4.498666 & 1919 \\
+MIN & 4.498666 & 1942 \\
+\bottomrule
+\end{longtable}
+
+NYY 1918: Babe Ruth got to the Yankees in 1920. And starting in 1919
+they had exactly one losing seasons between 1919 and 1965.\\
+BOS 1917, 1929, 1948, 1994: Their last world series win in the 1900s was
+in 1918.\\
+I don't know 1929. 1948 there was a big jump in runs?\\
+1994: Strike year.
+
+LAD: 1919, 1955
+
+ATL 1918 1949
+
+CHW: 1919: Blaack Sox Scandal
+
+CHC: 1919 1940 1940: War.
+
+Cin: 1929 1952
+
+CLE 1920 1939 Cleveland won the world series in 1920. Major change in
+offensive output.
+
+DET 1927 1940
+
+BAL 1919 1942
+
+SFG 1919 1937
+
+OAK 1920 1936 1951
+
+PHI 1918 1937
+
+PIT 1920 1946
+
+STL 1919 1941
+
+MIN 1919 1942
+
+Pearl Harbor was 1941. So US was in war in 1942.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-FortLee2007}{}}%
+Fort R., Lee Y. H. 2007. {``Structural Change, Competitive Balance, and
+the Rest of the Major Leagues.''} \emph{Economic Inquiry} 45 (3):
+519--32.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Groothius2017}{}}%
+Groothius, Peter A., Kurt W. Rotthoff, and Mark C. Strazicich. 2017.
+{``Structural Breaks in the Game: The Case of Major League Baseball.''}
+\emph{Journal of Sports Economics} 18 (6): 622--37.
+
+\leavevmode\vadjust pre{\hypertarget{ref-LeeFort2005}{}}%
+H., Lee Y., and Fort R. 2005. {``Structural Change in MLB Competitive
+Balance: The Depression, Team Location, and Integration.''}
+\emph{Economic Inquiry} 43 (1): 158--69.
+
+\leavevmode\vadjust pre{\hypertarget{ref-PalaciosHuerta2004}{}}%
+I., Palacios-Huerta. 2004. {``Structural Changes During a Century of the
+World's Most Popular Sport.''} \emph{Statistical Methods and
+Applications} 12: 241--58.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Nieswiadomy2012}{}}%
+Nieswiadomy, Strazicich, M. L. 2012. {``Was There a Structural Break in
+Barry Bonds's Bat?''} \emph{Journal of Quantitative Analysis in Sports}
+8 (3).
+
+\leavevmode\vadjust pre{\hypertarget{ref-BBHOF1869}{}}%
+Rothenberg, Matt. n.d. \emph{PRO BASEBALL BEGAN IN CINCINNATI IN 1869}.
+\url{https://baseballhall.org/discover/pro-baseball-began-in-cincinnati-in-1869\#:~:text=The\%20Cincinnati\%20Red\%20Stockings\%20made,professional\%20baseball\%20club\%20in\%201869.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Woltring2018}{}}%
+Woltring, Rost, M., and C. Jubenville. 2018. {``Examining Perceptions of
+Baseball's Eras: A Statistical Comparison.''} \emph{The Sport Journal}.
+\url{https://thesportjournal.org/article/examining-perceptions-of-baseballs-eras/}.
+
 \end{CSLReferences}

### 2023-04-06 manuscript/paper.Rmd
Added some intro and added a fun title.

@@
 title: |
-  |  \Large Eras in baseball: Change point analysis fun title. 
+  |  \Large Eras in baseball: Change-Ups and Change Points: An Exploration of Baseball's Historic Eras
 author: | 
@@
 
+More specifically, @Woltring2018 mentions six eras of modern baseball": 
+"Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras.
+A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005) (17)."  @Woltring2018 notes that they name a seventh era after 2006, which they term the Post Steroid Era.
+
 While many baseball writers have attempted to define the different eras of baseball, there has also been some academic work that has sought to empirically define eras in baseball.
@@
 @PalaciosHuerta2004 looked for structural changes in soccer using data from British soccer leagues through 1996.
-@FortLee2007 looked for structural breaks in major North American sports other than baseballl; that is basketball (NBA), hockey (NHL), and American football (NFL).
-WHAT DID THEY FIND?
+They had data from division I (Premier League) and II starting in the late 1800s, and data from lower divisions III and IV from just after WWII starting in 1947.
+They identify a number of change points.
+Notably they identify a change point in the mean of margin of victory in 1925 related to the change in the definition of offsides (changed from 3 players to 2 players), an change points in the variability of number of goals in the early 1980s and 1992 corresponding to the change in number of points for a win (i.e.
+went from 2 points to 3 points) and a change in the backpass rule, respectively.
+@FortLee2007 looked for structural breaks in major North American sports other than baseball (i.e.
+basketball (NBA), hockey (NHL), and American football (NFL)).
+They identified a number of change points related to competitive balance in each sport that often, though not always, correspond to league expansion, league mergers, or other major events in a sport (e.g.
+increased number of foreign players in the NBA in the late 1990s/early 2000s).
+
+<!-- WHAT DID THEY FIND?   ADD MORE HERE.  -->
 
@@
 First, we look for structural changes in the league as a whole across teams to empirically define different eras in baseball.
-Second, we look for change points within a team to determine eras within a team.
-In essence, this second goal can be used to identify the beginning and end of so called "dynasties", periods of sustained excellent performance by a team.
-In addition, we can also identify the opposite, sustained periods of poor performance which we will term ERAS of SUCKITUDE.
+Second, we look for change points within a team to determine eras of a team.
+This second analysis can be used to identify the beginning and end of so called "dynasties", periods of sustained excellent performance by a team.
+In addition, we can also identify the opposite, sustained periods of poor performance.
 
-Just Notes: 
-Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258
-@PalaciosHuerta2004 Structural Breaks in soccer
+<!-- Just Notes:  -->
+<!-- Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258 -->
+<!-- @PalaciosHuerta2004 Structural Breaks in soccer -->
 
-@Nieswiadomy2012 Was there a structural break in Barry Bonds’s Bat?
+<!-- @Nieswiadomy2012 Was there a structural break in Barry Bonds’s Bat?
 Looking for structural breaks in Bonds OPS times series.
@@
 The identify break points in June 1993 and Sept 2000.
+-->
 
-In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).
+<!-- In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014). -->
 
-https://onlinelibrary.wiley.com/doi/epdf/10.1111/j.1465-7295.2007.00026.x
-Fort and Lee 2007: Looks for structural breaks in NBA, NHL, and NFL
+<!-- https://onlinelibrary.wiley.com/doi/epdf/10.1111/j.1465-7295.2007.00026.x -->
+<!-- Fort and Lee 2007: Looks for structural breaks in NBA, NHL, and NFL -->
 
-https://www.baseball-reference.com/bullpen/History_of_baseball_in_the_United_States
+<!-- https://www.baseball-reference.com/bullpen/History_of_baseball_in_the_United_States -->
 
-https://thesportjournal.org/article/examining-perceptions-of-baseballs-eras/
+<!-- https://thesportjournal.org/article/examining-perceptions-of-baseballs-eras/ -->
 
@@
 
-https://pubmed.ncbi.nlm.nih.gov/30747582/?
+<!-- https://pubmed.ncbi.nlm.nih.gov/30747582/? -->
 
-The goal of this work is to use multivariate change point analysis to identify change points in MLB as a whole to objectively identify eras in the games history, but also to look for change points to identify team specific eras.
+<!-- The goal of this work is to use multivariate change point analysis to identify change points in MLB as a whole to objectively identify eras in the games history, but also to look for change points to identify team specific eras.
+-->
 
-The work that is the most closely related to our own is that of @Groothius2017, which looked for structural breaks (Note: the terms "structural breaks" and "change points" are used interchangeably in this manuscript) in univariate times series.
+<!-- The work that is the most closely related to our own is that of @Groothius2017, which looked for structural breaks (Note: the terms "structural breaks" and "change points" are used interchangeably in this manuscript) in univariate times series.
+-->
 
-univariate time series of the means and SD's of several statistics (i.e.
+<!-- univariate time series of the means and SD's of several statistics (i.e.
 SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrange Multiplier (LM) unit root test @LeeandStrazicich2003.
@@
 In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).
-
-@Woltring2018
-From Woltring:
-"Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras.
-A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005) (17).
-This study runs through the 2011 season and a seventh era will be added and labeled the Post Steroid Era (2006-2011)"
+-->
 

### 2023-05-19 manuscript/paper.Rmd
Some updates.

@@
   | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
+  | \normalsize Center for Data Science and Consulting \vspace{-1mm}
   | \normalsize Loyola University Chicago \vspace{-1mm}
@@
 
-Deadball era: https://sabr.org/journal/article/the-rise-and-fall-of-the-deadball-era/
+<!-- Deadball era: https://sabr.org/journal/article/the-rise-and-fall-of-the-deadball-era/ -->
 
-http://mlb.mlb.com/mlb/history/mlb_history_people.jsp?story=com
+<!-- http://mlb.mlb.com/mlb/history/mlb_history_people.jsp?story=com -->
 
-https://www.nytimes.com/2011/09/25/sports/baseball/scoring-in-baseball-returns-to-dead-ball-levels.html?_r=1&
+<!-- https://www.nytimes.com/2011/09/25/sports/baseball/scoring-in-baseball-returns-to-dead-ball-levels.html?_r=1& -->
 
-https://bleacherreport.com/articles/1676509-the-evolution-of-the-baseball-from-the-dead-ball-era-through-today
+<!-- https://bleacherreport.com/articles/1676509-the-evolution-of-the-baseball-from-the-dead-ball-era-through-today -->
 
-@FortLee2014: 
-Fort R., Lee Y. H. (2006). Stationarity and major league baseball attendance analysis. Journal of Sports Economics, 7, 408–415.
+<!-- @FortLee2006:  -->
+<!-- Fort R., Lee Y. H. (2006). Stationarity and major league baseball attendance analysis. Journal of Sports Economics, 7, 408–415. -->
 
-Fort R., Lee Y. H. (2007). Structural change, competitive balance, and the rest of the major leagues. Economic Inquiry, 45, 519–532.
+<!-- Fort R., Lee Y. H. (2007). Structural change, competitive balance, and the rest of the major leagues. Economic Inquiry, 45, 519–532. -->
 
-Lee Y. H., Fort R. (2005). Structural change in baseball’s competitive balance: The great depression, team location, and racial integration. Economic Inquiry, 43, 158–169.
+<!-- Lee Y. H., Fort R. (2005). Structural change in baseball’s competitive balance: The great depression, team location, and racial integration. Economic Inquiry, 43, 158–169. -->
 
-@Nieswiadomy2012 Was there a structural break in Barry Bonds’s Bat?
+<!-- @Nieswiadomy2012 Was there a structural break in Barry Bonds’s Bat?
 Looking for structural breaks in Bonds OPS times series.
@@
 The identify break points in June 1993 and Sept 2000.
+-->
 
-degruyter.com/document/doi/10.1515/1559-0410.1305/html
+<!-- degruyter.com/document/doi/10.1515/1559-0410.1305/html -->
 
-Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258
-@PalaciosHuerta2004 Structural Breaks in soccer
+<!-- Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258 -->
+<!-- @PalaciosHuerta2004 Structural Breaks in soccer -->
 
-@Groothius2017 is the most closely related work to ours.
+<!-- @Groothius2017 is the most closely related work to ours.
 They use univariate time series of the means and SD's of several statistics (i.e.
@@
 In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).
+-->
 
-Fort, Rodney and Young Hoon Lee (2006). “Stationarity and Major League Baseball Attendance Analysis.” Journal of Sports Economics, 7 (4), 408–415.
+<!-- Fort, Rodney and Young Hoon Lee (2006). “Stationarity and Major League Baseball Attendance Analysis.” Journal of Sports Economics, 7 (4), 408–415.  -->
 
-Fort, Rodney and Young Hoon Lee (2007). “Structural Change, Competitive Balance, and the Rest of the Major Leagues.” Economic Inquiry, 45 (3), 519-532.
+<!-- Fort, Rodney and Young Hoon Lee (2007). “Structural Change, Competitive Balance, and the Rest of the Major Leagues.” Economic Inquiry, 45 (3), 519-532.  -->
 
-Lee, Young Hoon, and Rodney Fort (2005). “Structural Change in Baseball’s Competitive Balance: The Great Depression, Team Location, and Racial Integration.” Economic Inquiry, 43, 158-169.
+<!-- Lee, Young Hoon, and Rodney Fort (2005). “Structural Change in Baseball’s Competitive Balance: The Great Depression, Team Location, and Racial Integration.” Economic Inquiry, 43, 158-169.  -->
 
-Sommers, P. (2008). “The Changing Hitting Performance Profile in Major League Baseball, 1966-2006.” Journal of Sports Economics, 9, 435-440.
+<!-- Sommers, P. (2008). “The Changing Hitting Performance Profile in Major League Baseball, 1966-2006.” Journal of Sports Economics, 9, 435-440.  -->
 
-@Berry1999bridging they talk about bridging eras.
+<!-- @Berry1999bridging they talk about bridging eras.   -->
 
-When did the steroids era start: https://www.espn.com/mlb/topics/_/page/the-steroids-era#:~:text=Unlike%20other%20MLB%20%22eras%2C%22,leaguewide%20PED%20testing%20until%202003.
+<!-- When did the steroids era start: https://www.espn.com/mlb/topics/_/page/the-steroids-era#:~:text=Unlike%20other%20MLB%20%22eras%2C%22,leaguewide%20PED%20testing%20until%202003. -->
 
-Traditional wisdom:
-1900-1919 dead ball era
+<!-- Traditional wisdom: -->
+<!-- 1900-1919 dead ball era -->
 
-From Woltring:
-"Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras.
+<!-- From Woltring: -->
+<!-- "Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras.
 A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005) (17).
-This study runs through the 2011 season and a seventh era will be added and labeled the Post Steroid Era (2006-2011)"
+This study runs through the 2011 season and a seventh era will be added and labeled the Post Steroid Era (2006-2011)" -->
 
-https://www.baseball-reference.com/bullpen/Deadball_Era
-1901-1920
+<!-- https://www.baseball-reference.com/bullpen/Deadball_Era -->
+<!-- 1901-1920 -->
 
-Mound was lowered in december 1968.  
-https://www.baseball-reference.com/bullpen/Pitcher%27s_mound
+<!-- Mound was lowered in december 1968.   -->
+<!-- https://www.baseball-reference.com/bullpen/Pitcher%27s_mound -->
 
-https://www.espn.com/mlb/story/_/id/33238595/major-league-baseball-stops-testing-players-steroids-nearly-20-years-report-says
+<!-- https://www.espn.com/mlb/story/_/id/33238595/major-league-baseball-stops-testing-players-steroids-nearly-20-years-report-says -->
 
-https://www.semanticscholar.org/paper/Was-There-a-Structural-Break-in-Barry-Bonds's-Bat-Nieswiadomy-Strazicich/0a32effa10a3d26ea4675efcca3f13229efe0f7d
+<!-- https://www.semanticscholar.org/paper/Was-There-a-Structural-Break-in-Barry-Bonds's-Bat-Nieswiadomy-Strazicich/0a32effa10a3d26ea4675efcca3f13229efe0f7d -->
 
@@
 
-teams <- read_csv("../data/Teams.csv") %>% 
+teams <- read_csv("/Users/gregorymatthews/Dropbox/baseball_cpt/data/Teams.csv") %>% 
   clean_names()
@@
 ## Results
-
 ```{r}
@@
 teams_cpt_time_df
-
 ```

### 2023-05-19 manuscript/paper.Rmd
Some updates.

@@
 "Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras.
-A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005) (17)."  @Woltring2018 notes that they name a seventh era after 2006, which they term the Post Steroid Era.
+A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005)."  @Woltring2018 notes that they name a seventh era after 2006, which they term the Post Steroid Era.
 
@@
 Use methods from Andrews1993, Bai1997, 1999 and Bai and Perron 1998 and 2003 to look for break points between 1901 -1999.
-Theuy measure competitive balance in two ways: 1) Noll 1988 and Scully 1989 and 2) Lee (2004).
+Theuy measure competitive balance in two ways: 1) @Noll1988 and @Scully1989 and 2) Lee 2004.
 They find break point sin competitive balance in 1912, 1926, and 1933 for the NL and in 1926 and 1957 in the AL.
@@
 -->
+
+<!-- https://journals.sagepub.com/doi/10.1177/152700250100200204 -->
 

### 2023-05-19 manuscript/paper.Rmd
Adding bib.
update

@@
 In addition, we can also identify the opposite, sustained periods of poor performance.
+
+@LeeFort2008: Attendance and the Uncertainty-of-Outcome Hypothesis in Baseball.
+Identify Break points in attendance in 1918 and 1945 for both leagues.
+For AL only: 1963, 1978, 1994.
+For NL only: 1976.
+Using Bai and Perron From table 3 in Mills and Fort 2014
+@MillsFort2014: LEAGUE-LEVEL ATTENDANCE AND OUTCOME UNCERTAINTY IN U.S.
+PRO SPORTS LEAGUES.
+Looks at NHL, NFL, NBA.
+Rottenburg 1956 looked at baseball.
+Identify break points in NBA, NFL, and NHL.
+Table 3
+@MillsSalaga2015: NCAA Basketball: League balance using stats.  
+@MillsFort2018: team-level: Attendence.  
+@SalagaFort2017: College football
 

### 2023-05-19 manuscript/paper.Rmd
updates.

@@
   | \normalsize [`mwhalen3@luc.edu`](mailto:mwhalen3@luc.edu) \vspace{-1mm}
+  |
+  | \large Brian M. Mills \vspace{-1.1mm}
+  | \normalsize College of Education \vspace{-1mm}
+  | \normalsize University of Texas at Austin \vspace{-1mm}
+  | \normalsize Austin, TX 60660 \vspace{-1mm}
+  | \normalsize [`brian.mills@austin.utexas.edu`](mailto:brian.mills@austin.utexas.edu) \vspace{-1mm}
   |

### 2023-07-14 manuscript/paper.Rmd
updating manuscript with data/methods section and few tables in the results

@@
 
+## Data
+
+We obtained multi-year baseball statistics from GREG??.
+Over the course of the sport's history, numerous statistics have been collected, and for our analysis, we focused on year-end statistics pertaining to the teams.
+To ensure consistency in our methodology, it was necessary for each team to have existed from 1900 until the end of the dataset in 2020.
+This timeframe of 120 years, denoted as T, encompassed a total of 16 franchise teams, namely NYY, BOS, LAD, ATL, CHW, CHC, CIN, CLE, DET, BAL, SFG, OAK, PHI, PIT, STL, and MIN.
+The statistics of interest for these teams include runs, hits, home runs, balls, strikeouts, at-bats, stolen bases, number of games played in a season, attendance, runs against, hits against, home runs against, balls against, and strikeouts against.
+However, a few statistics, such as balls, stolen bases, strikeouts, and attendance, had missing data for certain years.
+To address this, we employed multiple imputations using classification and regression trees (cite mice/cart) to fill in the missing values and ensure consistent data across all analyses.
+
+## Eras and Dynasties
+
+### Eras
+Since we are interested in finding where in time these different eras of baseball have occurred, we first examine each baseball statistic of interest as their own analysis for all 16 teams.
+Those baseball statistics are home runs, strikeouts, balls, stolen bases, runs, and attendance which were all standardized to the number of games played in a season since we are interested in seeing how the different statistics change from season to season.
+Using the methodology from Cho and Fry (CITE) a thresholding parameter was determined from bootstrapping and then used to find where change points in mean and variance existed in the panel of teams data of a given baseball statistic.
+To examine the idea of changes throughout teams and types of statistics (initially), for each season the average baseball statistic was found from all the teams that year and then an average time series was created of each of the types of statistics of interest.
+Then the change point methodology was performed on those 6 time series of average statistic types.
+
+###Dynasties
+
+Continuing our investigation, we explored the concept of "dynasties" within modern baseball teams, irrespective of their length of a teams existence.
+A dynasty, in this context, refers to a period characterized by significant success or potentially hardships.
+Such a dynasty is identified when a team experiences a noticeable change across all their statistics in a given season, whether it be in a positive or negative direction.
+This approach aligns well with our methodology, as it involves examining panels of time series data encompassing all statistics for a singular team.
+The statistics of interest for this analysis include runs, hits, home runs, balls, strikeouts, runs against, hits against, home runs against, balls against, and strikeouts against.
+To ensure comparability, these statistics were standardized using the season average and standard deviation derived from all teams present in each respective season.
+By standardizing the statistics, we capture a team's offensive and defensive performance.
+Notably, if a team exhibits a high and positive offensive performance, it can be balanced to zero by large and negative defensive statistics when standardized.
+Therefore, if a team excels in both aspects of the game, it would have numerous time points above zero, whereas underperforming teams would display many below-zero values.For each team, we examined their ten statistics, analyzing both mean and variance changes to identify potential change points.
+By scrutinizing these indicators, we aimed to uncover shifts in a team's performance and assess their significance within the context of dynasties.
+
 # Results
-```{r echo = FALSE}
-library(tidyverse)
-library(janitor)
-library(hdbinseg)
 
-teams <- read_csv("/Users/gregorymatthews/Dropbox/baseball_cpt/data/Teams.csv") %>% 
-  clean_names()
+## Tables
 
-team_names <- teams %>% 
-  group_by(franch_id) %>% 
-  count() %>% 
-  filter(n > 120) %>% 
-  pull(franch_id)
+|Mean/Variance CPT     |Franchise | Change Point Year |
+|:--------|:---------|:----------------|
+|mean     |ANA       |1977             |
+|mean     |ATL       |1900, 1990       |
+|mean     |BAL       |1959, 1999       |
+|mean     |BOS       |1918, 1937       |
+|mean     |CHC       |1891             |
+|mean     |CHW       |1970             |
+|mean     |CIN       |1918, 1952       |
+|mean     |CLE       |1944, 1959, 1993 |
+|mean     |DET       |1958, 1988       |
+|mean     |FLA       |2011             |
+|mean     |HOU       |1999             |
+|mean     |KCR       |1997             |
+|mean     |LAD       |1940, 1961       |
+|mean     |MIL       |1977, 1996       |
+|mean     |MIN       |1959             |
+|mean     |PHI       |1917, 1949       |
+|mean     |PIT       |1939             |
+|mean     |SDP       |1977             |
+|mean     |SEA       |1986, 1999       |
+|mean     |SFG       |1903, 1973       |
+|mean     |STL       |1937             |
+|mean     |TBD       |2007             |
+|mean     |TEX       |1973, 1990       |
+|mean     |WSN       |2011             |
+|variance |ATL       |1887             |
+|variance |BOS       |2008             |
+|variance |CHW       |1969             |
+|variance |CLE       |1966, 1981, 1993 |
+|variance |DET       |1942             |
+|variance |HOU       |2010             |
+|variance |NYM       |1974             |
+|variance |PIT       |1902, 2001       |
+|variance |SDP       |1977             |
+|variance |SFG       |1918             |
 
-# teams %>% 
-#   filter(franch_id %in% team_names) %>% 
-#   group_by(franch_id) %>% 
-#   summarise(min(year_id)) ## filter after 1901
-
-old_teams_full <- teams %>% 
-  filter(franch_id %in% team_names & year_id > 1900) %>% 
-  select(year_id, franch_id, w, l, r, ab, h, hr, bb, so)
-
-teams_wide <- old_teams_full %>% 
-  pivot_wider(id_cols = franch_id, 
-              names_from = year_id, 
-              values_from = hr)
-
-## find thresholding value for cpt using bootstrapping
-thrs <- dcbs.thr(as.matrix(teams_wide[-1]), interval = c(1, dim(teams_wide[-1])[2]))
-### if I lower the threshold more changepoints
-hr_cpt_dcbs <- dcbs.alg(as.matrix(teams_wide[-1]), cp.type = 1, thr = thrs-1)
-hr_cpt_dcbs$ecp
-hr_cpt_sbs <- sbs.alg(as.matrix(teams_wide[-1]), cp.type = 1)         
-hr_cpt_sbs$ecp
-
-cpt_years <- as.numeric(names(teams_wide[-1])[hr_cpt_dcbs$ecp])
-
-ggplot(old_teams_full, aes(year_id, hr, color = franch_id)) +
-  geom_line() +
-  geom_vline(xintercept = cpt_years)
-
-```
-
-## Results
-```{r}
-load("/Users/gregorymatthews/Dropbox/baseball_cpt/data/teams_cpt_time_df.RData")
-teams_cpt_time_df
-```
+|Mean/Variance CPT   |Baseball Statistic     | Change Point Year                    |
+|:--------|:------------|:----------------------|
+|mean     |att_per_game |1945, 1975, 1992       |
+|mean     |bb_per_game  |1927                   |
+|mean     |hr_per_game  |1920, 1946, 1967, 1993 |
+|mean     |r_per_game   |1919, 1940             |
+|mean     |sb_per_game  |1919                   |
+|mean     |so_per_game  |1956, 1993, 2009       |
+|variance |att_per_game |1944, 2008             |
+|variance |hr_per_game  |1927, 1944, 1995, 2007 |
+|variance |r_per_game   |1935                   |
+|variance |sb_per_game  |1919, 1966             |
+|variance |so_per_game  |1990                   |
 

### 2023-07-14 manuscript/paper.tex
updating manuscript with data/methods section and few tables in the results

--- 2023-02-18 manuscript/paper.tex
@@
 ]{article}
-\title{~\Large Eras in baseball: Change point analysis fun title.}
-\author{\large Mena Whalen \vspace{-1.1mm}\\
-\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
-\normalsize Loyola University Chicago \vspace{-1mm}\\
-\normalsize Chicago, IL 60660 \vspace{-1mm}\\
-\normalsize \href{mailto:mwhalen3@luc.edu}{\texttt{mwhalen3@luc.edu}}
-\vspace{-1mm}\\
-\strut \\
-\large Gregory J. Matthews \vspace{-1.1mm}\\
-\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
-\normalsize Loyola University Chicago \vspace{-1mm}\\
-\normalsize Chicago, IL 60660 \vspace{-1mm}\\
-\normalsize \href{mailto:gmatthews1@luc.edu}{\texttt{gmatthews1@luc.edu}}
-\vspace{-1mm}}
-\date{}
-
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
 \fi
@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  colorlinks=true,
-  linkcolor={cyan},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={cyan},
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
@@
 \fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same}
+\hypersetup{
+  colorlinks=true,
+  linkcolor={cyan},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={cyan},
+  pdfcreator={LaTeX via pandoc}}
+
+\title{~\Large Eras in baseball: Change-Ups and Change Points: An
+Exploration of Baseball's Historic Eras}
+\author{\large Mena Whalen \vspace{-1.1mm}\\
+\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
+\normalsize Center for Data Science and Consulting \vspace{-1mm}\\
+\normalsize Loyola University Chicago \vspace{-1mm}\\
+\normalsize Chicago, IL 60660 \vspace{-1mm}\\
+\normalsize \href{mailto:mwhalen3@luc.edu}{\texttt{mwhalen3@luc.edu}}
+\vspace{-1mm}\\
+\strut \\
+\large Brian M. Mills \vspace{-1.1mm}\\
+\normalsize College of Education \vspace{-1mm}\\
+\normalsize University of Texas at Austin \vspace{-1mm}\\
+\normalsize Austin, TX 60660 \vspace{-1mm}\\
+\normalsize \href{mailto:brian.mills@austin.utexas.edu}{\texttt{brian.mills@austin.utexas.edu}}
+\vspace{-1mm}\\
+\strut \\
+\large Gregory J. Matthews \vspace{-1.1mm}\\
+\normalsize Department of Mathematics and Statistics \vspace{-1mm}\\
+\normalsize Center for Data Science and Consulting \vspace{-1mm}\\
+\normalsize Loyola University Chicago \vspace{-1mm}\\
+\normalsize Chicago, IL 60660 \vspace{-1mm}\\
+\normalsize \href{mailto:gmatthews1@luc.edu}{\texttt{gmatthews1@luc.edu}}
+\vspace{-1mm}}
+\date{}
 
@@
 
-\protect\hyperlink{ref-Berry1999bridging}{Berry, Reese, and Larkey}
-(\protect\hyperlink{ref-Berry1999bridging}{1999}) they talk about
-bridging eras.
+The first professional baseball team in the United States, the
+Cinicnnati Red Stockings, was formed in 1869 (Rothenberg
+(\protect\hyperlink{ref-BBHOF1869}{n.d.})). Many leagues came and went
+in the late 1800s, but National League (NL), formed in 1876, emerged as
+the predominant league of the time. A few decades later, the American
+League (AL) began growing in popularity and eventually reached an
+agreement with the NL to be the two major leagues of baseball with the
+winner of each league playing in the World Series starting in 1903.
 
-When did the steroids era start:
-\url{https://www.espn.com/mlb/topics/_/page/the-steroids-era\#}:\textasciitilde:text=Unlike\%20other\%20MLB\%20\%22eras\%2C\%22,leaguewide\%20PED\%20testing\%20until\%202003.
+Throughout the history of baseball in the United States, the game has
+gone through many changes and distinct eras. For example, the time
+period between approximately 1900-1919 is often referred to as the
+``Dead Ball Era'' and was marked by low scoring games and dominant
+pitching. Another more recent example would be the ``Steroid Era'' which
+lasted from approximately 1994 through 2005 and was characterized by a
+rapid increase in power hitting largely attributed to players using
+performance enhancing drugs.
 
-Traditional wisdom: 1900-1919 dead ball era
-
-From Woltring: ``Baseball has endured much change over the course of its
+More specifically, Woltring and Jubenville
+(\protect\hyperlink{ref-Woltring2018}{2018}) mentions six eras of modern
+baseball'': ``Baseball has endured much change over the course of its
 history, and because of constant change, the modern era of baseball has
@@
 Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long
-Ball/Steroid Era (1994-2005) (17). This study runs through the 2011
-season and a seventh era will be added and labeled the Post Steroid Era
-(2006-2011)''
+Ball/Steroid Era (1994-2005).'' Woltring and Jubenville
+(\protect\hyperlink{ref-Woltring2018}{2018}) notes that they name a
+seventh era after 2006, which they term the Post Steroid Era.
 
-\url{https://www.baseball-reference.com/bullpen/Deadball_Era} 1901-1920
+While many baseball writers have attempted to define the different eras
+of baseball, there has also been some academic work that has sought to
+empirically define eras in baseball. Groothius, Rotthoff, and Strazicich
+(\protect\hyperlink{ref-Groothius2017}{2017}) looked for structural
+breaks in univariate times series of performance measures over the
+period from 1871-2020. They analyzed four statistics: slugging
+percentage (SLG, home run (HR) rate, batting average (BA), and runs
+batted in (RBI) rate. For each of these statistics, the computed the
+mean and standard deviation (SD) across all players who had at least 100
+at bats in a given season to yield a univariate time series for each of
+these statistics. They then used the Lagrange Multiplier (LM) unit root
+test proposed in J. Lee and Strazicich
+(\protect\hyperlink{ref-LeeandStrazicich2003}{n.d.}) to find structural
+breaks. They identified structural breaks in slugging percentage in 1921
+and 1992, the first of which marks the end of the Dead Ball Era and the
+latter corresponding to the start of the steroid era.
 
-Mound was lowered in december 1968.\\
-\url{https://www.baseball-reference.com/bullpen/Pitcher\%27s_mound}
+Lee and Fort (\protect\hyperlink{ref-LeeFort2005}{2005}) looks for
+structural changes in competitive balance of the two league American and
+National. Use methods from Andrews1993, Bai1997, 1999 and Bai and Perron
+1998 and 2003 to look for break points between 1901 -1999. Theuy measure
+competitive balance in two ways: 1) Noll
+(\protect\hyperlink{ref-Noll1988}{1988}) and Scully
+(\protect\hyperlink{ref-Scully1989}{1989}) and 2) Lee 2004. They find
+break point sin competitive balance in 1912, 1926, and 1933 for the NL
+and in 1926 and 1957 in the AL.
 
-\url{https://www.espn.com/mlb/story/_/id/33238595/major-league-baseball-stops-testing-players-steroids-nearly-20-years-report-says}
+Baseball is not the only sport where this type of analysis has been
+applied. I. (\protect\hyperlink{ref-PalaciosHuerta2004}{2004}) looked
+for structural changes in soccer using data from British soccer leagues
+through 1996. They had data from division I (Premier League) and II
+starting in the late 1800s, and data from lower divisions III and IV
+from just after WWII starting in 1947. They identify a number of change
+points. Notably they identify a change point in the mean of margin of
+victory in 1925 related to the change in the definition of offsides
+(changed from 3 players to 2 players), an change points in the
+variability of number of goals in the early 1980s and 1992 corresponding
+to the change in number of points for a win (i.e.~went from 2 points to
+3 points) and a change in the backpass rule, respectively. Fort and Lee
+(\protect\hyperlink{ref-FortLee2007}{2007}) looked for structural breaks
+in major North American sports other than baseball (i.e.~basketball
+(NBA), hockey (NHL), and American football (NFL)). They identified a
+number of change points related to competitive balance in each sport
+that often, though not always, correspond to league expansion, league
+mergers, or other major events in a sport (e.g.~increased number of
+foreign players in the NBA in the late 1990s/early 2000s).
+
+All of the previous work mentioned here focuses on change point analysis
+in \textbf{\emph{univariate}} time series. However, recent methodolgical
+developments in change point analysis allow for change point analysis in
+\textbf{\emph{multivariate}} time series, which is the focus of this
+current manuscript. This work leverages techniques such as the Double
+CUSUM Binary Segmentation algorithm (H. Cho
+(\protect\hyperlink{ref-Cho2016}{2016})) and the Sparsified Binary
+Segmentation algorithm (H. Cho and Fryzlewicz
+(\protect\hyperlink{ref-ChoFryzlewwicz2014}{2014})) to look for change
+points in Major League Baseball at two main levels. First, we look for
+structural changes in the league as a whole across teams to empirically
+define different eras in baseball. Second, we look for change points
+within a team to determine eras of a team. This second analysis can be
+used to identify the beginning and end of so called ``dynasties'',
+periods of sustained excellent performance by a team. In addition, we
+can also identify the opposite, sustained periods of poor performance.
+
+Y. H. Lee and Fort (\protect\hyperlink{ref-LeeFort2008}{2008}):
+Attendance and the Uncertainty-of-Outcome Hypothesis in Baseball.
+Identify Break points in attendance in 1918 and 1945 for both leagues.
+For AL only: 1963, 1978, 1994. For NL only: 1976. Using Bai and Perron
+From table 3 in Mills and Fort 2014 MILLS and FORT
+(\protect\hyperlink{ref-MillsFort2014}{2014}): LEAGUE-LEVEL ATTENDANCE
+AND OUTCOME UNCERTAINTY IN U.S. PRO SPORTS LEAGUES. Looks at NHL, NFL,
+NBA. Rottenburg 1956 looked at baseball. Identify break points in NBA,
+NFL, and NHL. Table 3 Mills and Salaga
+(\protect\hyperlink{ref-MillsSalaga2015}{2015}): NCAA Basketball: League
+balance using stats.\\
+Mills and Fort (\protect\hyperlink{ref-MillsFort2018}{2018}):
+team-level: Attendence.\\
+Salaga and Fort (\protect\hyperlink{ref-SalagaFort2017}{2017}): College
+football
+
+SOME MORE STUFF
 
@@
 
-\protect\hyperlink{ref-ChoFryzlewwicz2014}{H. Cho and Fryzlewicz}
-(\protect\hyperlink{ref-ChoFryzlewwicz2014}{2014}) and
-\protect\hyperlink{ref-Cho2016}{H. Cho}
-(\protect\hyperlink{ref-Cho2016}{2016})
+H. Cho and Fryzlewicz (\protect\hyperlink{ref-ChoFryzlewwicz2014}{2014})
+and H. Cho (\protect\hyperlink{ref-Cho2016}{2016})
 
-R Pacakge: \protect\hyperlink{ref-hdbinseg}{Haeran Cho and Fryzlewicz}
+R Pacakge: Haeran Cho and Fryzlewicz
 (\protect\hyperlink{ref-hdbinseg}{2018})
+
+\hypertarget{data}{%
+\subsection{Data}\label{data}}
+
+We obtained multi-year baseball statistics from GREG??. Over the course
+of the sport's history, numerous statistics have been collected, and for
+our analysis, we focused on year-end statistics pertaining to the teams.
+To ensure consistency in our methodology, it was necessary for each team
+to have existed from 1900 until the end of the dataset in 2020. This
+timeframe of 120 years, denoted as T, encompassed a total of 16
+franchise teams, namely NYY, BOS, LAD, ATL, CHW, CHC, CIN, CLE, DET,
+BAL, SFG, OAK, PHI, PIT, STL, and MIN. The statistics of interest for
+these teams include runs, hits, home runs, balls, strikeouts, at-bats,
+stolen bases, number of games played in a season, attendance, runs
+against, hits against, home runs against, balls against, and strikeouts
+against. However, a few statistics, such as balls, stolen bases,
+strikeouts, and attendance, had missing data for certain years. To
+address this, we employed multiple imputations using classification and
+regression trees (cite mice/cart) to fill in the missing values and
+ensure consistent data across all analyses.
+
+\hypertarget{eras-and-dynasties}{%
+\subsection{Eras and Dynasties}\label{eras-and-dynasties}}
+
+\hypertarget{eras}{%
+\subsubsection{Eras}\label{eras}}
+
+Since we are interested in finding where in time these different eras of
+baseball have occurred, we first examine each baseball statistic of
+interest as their own analysis for all 16 teams. Those baseball
+statistics are home runs, strikeouts, balls, stolen bases, runs, and
+attendance which were all standardized to the number of games played in
+a season since we are interested in seeing how the different statistics
+change from season to season. Using the methodology from Cho and Fry
+(CITE) a thresholding parameter was determined from bootstrapping and
+then used to find where change points in mean and variance existed in
+the panel of teams data of a given baseball statistic. To examine the
+idea of changes throughout teams and types of statistics (initially),
+for each season the average baseball statistic was found from all the
+teams that year and then an average time series was created of each of
+the types of statistics of interest. Then the change point methodology
+was performed on those 6 time series of average statistic types.
+
+\#\#\#Dynasties
+
+Continuing our investigation, we explored the concept of ``dynasties''
+within modern baseball teams, irrespective of their length of a teams
+existence. A dynasty, in this context, refers to a period characterized
+by significant success or potentially hardships. Such a dynasty is
+identified when a team experiences a noticeable change across all their
+statistics in a given season, whether it be in a positive or negative
+direction. This approach aligns well with our methodology, as it
+involves examining panels of time series data encompassing all
+statistics for a singular team. The statistics of interest for this
+analysis include runs, hits, home runs, balls, strikeouts, runs against,
+hits against, home runs against, balls against, and strikeouts against.
+To ensure comparability, these statistics were standardized using the
+season average and standard deviation derived from all teams present in
+each respective season. By standardizing the statistics, we capture a
+team's offensive and defensive performance. Notably, if a team exhibits
+a high and positive offensive performance, it can be balanced to zero by
+large and negative defensive statistics when standardized. Therefore, if
+a team excels in both aspects of the game, it would have numerous time
+points above zero, whereas underperforming teams would display many
+below-zero values.For each team, we examined their ten statistics,
+analyzing both mean and variance changes to identify potential change
+points. By scrutinizing these indicators, we aimed to uncover shifts in
+a team's performance and assess their significance within the context of
+dynasties.
 
@@
 
-\begin{verbatim}
-## [1] 20 46 67 94
-\end{verbatim}
+\hypertarget{tables}{%
+\subsection{Tables}\label{tables}}
 
-\begin{verbatim}
-##  b 
-## 48
-\end{verbatim}
+\begin{longtable}[]{@{}lll@{}}
+\toprule\noalign{}
+Mean/Variance CPT & Franchise & Change Point Year \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+mean & ANA & 1977 \\
+mean & ATL & 1900, 1990 \\
+mean & BAL & 1959, 1999 \\
+mean & BOS & 1918, 1937 \\
+mean & CHC & 1891 \\
+mean & CHW & 1970 \\
+mean & CIN & 1918, 1952 \\
+mean & CLE & 1944, 1959, 1993 \\
+mean & DET & 1958, 1988 \\
+mean & FLA & 2011 \\
+mean & HOU & 1999 \\
+mean & KCR & 1997 \\
+mean & LAD & 1940, 1961 \\
+mean & MIL & 1977, 1996 \\
+mean & MIN & 1959 \\
+mean & PHI & 1917, 1949 \\
+mean & PIT & 1939 \\
+mean & SDP & 1977 \\
+mean & SEA & 1986, 1999 \\
+mean & SFG & 1903, 1973 \\
+mean & STL & 1937 \\
+mean & TBD & 2007 \\
+mean & TEX & 1973, 1990 \\
+mean & WSN & 2011 \\
+variance & ATL & 1887 \\
+variance & BOS & 2008 \\
+variance & CHW & 1969 \\
+variance & CLE & 1966, 1981, 1993 \\
+variance & DET & 1942 \\
+variance & HOU & 2010 \\
+variance & NYM & 1974 \\
+variance & PIT & 1902, 2001 \\
+variance & SDP & 1977 \\
+variance & SFG & 1918 \\
+\end{longtable}
 
-\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-1-1} \end{center}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2000}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2889}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5111}}@{}}
+\toprule\noalign{}
+\begin{minipage}[b]{\linewidth}\raggedright
+Mean/Variance CPT
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Baseball Statistic
+\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
+Change Point Year
+\end{minipage} \\
+\midrule\noalign{}
+\endhead
+\bottomrule\noalign{}
+\endlastfoot
+mean & att\_per\_game & 1945, 1975, 1992 \\
+mean & bb\_per\_game & 1927 \\
+mean & hr\_per\_game & 1920, 1946, 1967, 1993 \\
+mean & r\_per\_game & 1919, 1940 \\
+mean & sb\_per\_game & 1919 \\
+mean & so\_per\_game & 1956, 1993, 2009 \\
+variance & att\_per\_game & 1944, 2008 \\
+variance & hr\_per\_game & 1927, 1944, 1995, 2007 \\
+variance & r\_per\_game & 1935 \\
+variance & sb\_per\_game & 1919, 1966 \\
+variance & so\_per\_game & 1990 \\
+\end{longtable}
+
+NYY 1918: Babe Ruth got to the Yankees in 1920. And starting in 1919
+they had exactly one losing seasons between 1919 and 1965.\\
+BOS 1917, 1929, 1948, 1994: Their last world series win in the 1900s was
+in 1918.\\
+I don't know 1929. 1948 there was a big jump in runs?\\
+1994: Strike year.
+
+LAD: 1919, 1955
+
+ATL 1918 1949
+
+CHW: 1919: Blaack Sox Scandal
+
+CHC: 1919 1940 1940: War.
+
+Cin: 1929 1952
+
+CLE 1920 1939 Cleveland won the world series in 1920. Major change in
+offensive output.
+
+DET 1927 1940
+
+BAL 1919 1942
+
+SFG 1919 1937
+
+OAK 1920 1936 1951
+
+PHI 1918 1937
+
+PIT 1920 1946
+
+STL 1919 1941
+
+MIN 1919 1942
+
+Pearl Harbor was 1941. So US was in war in 1942.
 
@@
 We thank Michael Lopez for suggesting we do ``something with change
-point analysis.''
+point analysis''.
 
@@
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-Berry1999bridging}{}}%
-Berry, Scott M., C. Shane Reese, and Patrick D. Larkey. 1999.
-{``Bridging Different Eras in Sports.''} \emph{Journal of the American
-Statistical Association} 94 (447): 661--76.
-\url{https://doi.org/10.1080/01621459.1999.10474163}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-Cho2016}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-FortLee2007}{}}%
+Fort, and Lee. 2007. {``Structural Change, Competitive Balance, and the
+Rest of the Major Leagues.''} \emph{Economic Inquiry} 45 (3): 519--32.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Groothius2017}{}}%
+Groothius, Peter A., Kurt W. Rotthoff, and Mark C. Strazicich. 2017.
+{``Structural Breaks in the Game: The Case of Major League Baseball.''}
+\emph{Journal of Sports Economics} 18 (6): 622--37.
+
+\leavevmode\vadjust pre{\hypertarget{ref-PalaciosHuerta2004}{}}%
+I., Palacios-Huerta. 2004. {``Structural Changes During a Century of the
+World's Most Popular Sport.''} \emph{Statistical Methods and
+Applications} 12: 241--58.
+
+\leavevmode\vadjust pre{\hypertarget{ref-LeeandStrazicich2003}{}}%
+Lee, J., and M. Strazicich. n.d. {``Minimum Lagrange Multiplier Unit
+Root Test with Two Structural Breaks.''} \emph{The Review of Economics
+and Statistics}, no. 4: 1082--89.
+
+\leavevmode\vadjust pre{\hypertarget{ref-LeeFort2008}{}}%
+Lee, Y. H., and R. Fort. 2008. {``Attendance and the
+Uncertainty-of-Outcome Hypothesis in Baseball.''} \emph{Review of
+Industrial Organization} 33 (4): 281--95.
+
+\leavevmode\vadjust pre{\hypertarget{ref-LeeFort2005}{}}%
+Lee, and Fort. 2005. {``Structural Change in MLB Competitive Balance:
+The Depression, Team Location, and Integration.''} \emph{Economic
+Inquiry} 43 (1): 158--69.
+
+\leavevmode\vadjust pre{\hypertarget{ref-MillsFort2018}{}}%
+Mills, Brian M., and Rodney Fort. 2018. {``Team-Level Time Series
+Analysis in MLB, the NBA, and the NHL: Attendance and Outcome
+Uncertainty.''} \emph{Journal of Sports Economics} 19 (7): 911--33.
+\url{https://doi.org/10.1177/1527002517690787}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-MillsSalaga2015}{}}%
+Mills, Brian M., and Steven Salaga. 2015. {``Historical Time Series
+Perspectives on Competitive Balance in NCAA Division i Basketball.''}
+\emph{Journal of Sports Economics} 16 (6): 614--46.
+\url{https://doi.org/10.1177/1527002515580925}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-MillsFort2014}{}}%
+MILLS, BRIAN, and RODNEY FORT. 2014. {``LEAGUE-LEVEL ATTENDANCE AND
+OUTCOME UNCERTAINTY IN u.s. PRO SPORTS LEAGUES.''} \emph{Economic
+Inquiry} 52 (1): 205--18.
+https://doi.org/\url{https://doi.org/10.1111/ecin.12037}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Noll1988}{}}%
+Noll, R. G. 1988. {``Professional Basketball.''} \emph{Stanford
+University Studies in Industrial Economics}, no. 144.
+
+\leavevmode\vadjust pre{\hypertarget{ref-BBHOF1869}{}}%
+Rothenberg, Matt. n.d. \emph{PRO BASEBALL BEGAN IN CINCINNATI IN 1869}.
+\url{https://baseballhall.org/discover/pro-baseball-began-in-cincinnati-in-1869\#:~:text=The\%20Cincinnati\%20Red\%20Stockings\%20made,professional\%20baseball\%20club\%20in\%201869.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-SalagaFort2017}{}}%
+Salaga, S., and R. Fort. 2017. {``Structural Change in Competitive
+Balance in Big-Time College Football.''} \emph{Review of Industrial
+Organization} 50: 27--41.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Scully1989}{}}%
+Scully, G. W. 1989. \emph{The Business of Major League Baseball}.
+University of Chicago Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Woltring2018}{}}%
+Woltring, Rost, M., and C. Jubenville. 2018. {``Examining Perceptions of
+Baseball's Eras: A Statistical Comparison.''} \emph{The Sport Journal}.
+\url{https://thesportjournal.org/article/examining-perceptions-of-baseballs-eras/}.
+
 \end{CSLReferences}

### 2023-07-14 cpts_all_variables.R

@@ -116,6 +116,11 @@ full_teams %>%
   geom_vline(data = bind_rows(time_cpts, time_cpts_var), aes(xintercept = years, linetype = change)) +
   facet_wrap(~ variable, scales = "free_y")
 
+bind_rows(time_cpts, time_cpts_var) %>% 
+  group_by(change, variable) %>% 
+  summarise(cpt = paste(years, collapse = ", ")) %>% 
+  kableExtra::kable(format = "markdown")
+
 ##### look at aggregate of a season (year) of the teams to look at all the stats together
 
 szn_ave <- full_teams %>% 

### 2023-07-14 current_teams_cpts.R

@@ -109,3 +109,10 @@ std_teams %>%
   geom_line() +
   geom_vline(data = total_values, aes(xintercept = years, linetype = type)) +
   facet_wrap(~franch_id, scales = "free_x")
+
+## print out table of results
+total_values %>% 
+  group_by(type, franch_id) %>% 
+  summarise(cpt = paste(years, collapse = ", ")) %>% 
+  kableExtra::kable(format = "markdown")
+

### 2023-07-19 manuscript/paper.Rmd
Adding methods section

@@
 
+A commonly employed method for detecting changes over time involves the CUSUM statistic in binary segmentation, which creates a tree diagram, dividing the time series after successfully identifying a change point until no further change points are detected.
+The magnitude of the CUSUM difference between two segments generally indicates a potential change point, depending on the assumptions and statistical tests applied to the time series.
+While this process works well for a single time series, detecting change points across a panel of multiple time series requires a more comprehensive approach.
+@Cho2016 introduced the double CUSUM (DC) algorithm, assuming there are $n$ time series, all of the same length $T$, with unknown common change point locations.
+This algorithm utilizes the CUSUM values of multiple time series $(j = 1, \dots, n)$ to cross-sectionally compare time segments for potential change points.
+
+To achieve this, Cho's DC statistic involves ordered CUSUM values for a given time location, favoring larger values as potential change point candidates.
+It incorporates a partial thresholding of the smaller ordered CUSUM values, effectively representing a scaled average of the most prominent values at a given location across all "n" time series.
+Maximizing the DC statistic across all time series and potential time locations results in a test statistic for detecting a change point.
+This test statistic is then compared against a specific testing criterion, denoted as $\pi_{n,T}^\phi$, leading to the partitioning of time series and the formation of a tree structure based on the detected change points.
+
+To determine the testing criterion $\pi_{n,T}^\phi$, a Generalized Dynamic Factor Model (GDFM) bootstrapping algorithm is employed, accounting for potential correlations within and between high-dimensional time series.
+This methodology enables the detection of second-order change points, utilizing methodology from @ChoFryzlewwicz2014 Haar wavelet periodograms and cross-periodograms, as opposed to the traditional CUSUM statistic.
+For further details, refer to @Cho2016 and @ChoFryzlewwicz2014.
+
 ## Data

### 2023-07-19 manuscript/paper.tex
Adding methods section

@@
 (\protect\hyperlink{ref-hdbinseg}{2018})
+
+A commonly employed method for detecting changes over time involves the
+CUSUM statistic in binary segmentation, which creates a tree diagram,
+dividing the time series after successfully identifying a change point
+until no further change points are detected. The magnitude of the CUSUM
+difference between two segments generally indicates a potential change
+point, depending on the assumptions and statistical tests applied to the
+time series. While this process works well for a single time series,
+detecting change points across a panel of multiple time series requires
+a more comprehensive approach. H. Cho
+(\protect\hyperlink{ref-Cho2016}{2016}) introduced the double CUSUM (DC)
+algorithm, assuming there are \(n\) time series, all of the same length
+\(T\), with unknown common change point locations. This algorithm
+utilizes the CUSUM values of multiple time series \((j = 1, \dots, n)\)
+to cross-sectionally compare time segments for potential change points.
+
+To achieve this, Cho's DC statistic involves ordered CUSUM values for a
+given time location, favoring larger values as potential change point
+candidates. It incorporates a partial thresholding of the smaller
+ordered CUSUM values, effectively representing a scaled average of the
+most prominent values at a given location across all ``n'' time series.
+Maximizing the DC statistic across all time series and potential time
+locations results in a test statistic for detecting a change point. This
+test statistic is then compared against a specific testing criterion,
+denoted as \(\pi_{n,T}^\phi\), leading to the partitioning of time
+series and the formation of a tree structure based on the detected
+change points.
+
+To determine the testing criterion \(\pi_{n,T}^\phi\), a Generalized
+Dynamic Factor Model (GDFM) bootstrapping algorithm is employed,
+accounting for potential correlations within and between
+high-dimensional time series. This methodology enables the detection of
+second-order change points, utilizing methodology from H. Cho and
+Fryzlewicz (\protect\hyperlink{ref-ChoFryzlewwicz2014}{2014}) Haar
+wavelet periodograms and cross-periodograms, as opposed to the
+traditional CUSUM statistic. For further details, refer to H. Cho
+(\protect\hyperlink{ref-Cho2016}{2016}) and H. Cho and Fryzlewicz
+(\protect\hyperlink{ref-ChoFryzlewwicz2014}{2014}).
 

### 2023-07-19 manuscript/paper.Rmd
updated.

--- 2023-05-19 manuscript/paper.Rmd
@@
   | \normalsize University of Texas at Austin \vspace{-1mm}
-  | \normalsize Austin, TX 60660 \vspace{-1mm}
+  | \normalsize Austin, TX 78712 \vspace{-1mm}
   | \normalsize [`brian.mills@austin.utexas.edu`](mailto:brian.mills@austin.utexas.edu) \vspace{-1mm}

### 2023-07-19 manuscript/paper.tex
updated.

--- 2023-07-14 manuscript/paper.tex
@@
 \usepackage{amsmath,amssymb}
+\usepackage{lmodern}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math} % this also loads fontspec
+  \usepackage{unicode-math}
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else
-  % xetex/luatex font selection
 \fi
@@
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same}
+\urlstyle{same} % disable monospaced font for URLs
 \hypersetup{
@@
 \normalsize University of Texas at Austin \vspace{-1mm}\\
-\normalsize Austin, TX 60660 \vspace{-1mm}\\
+\normalsize Austin, TX 78712 \vspace{-1mm}\\
 \normalsize \href{mailto:brian.mills@austin.utexas.edu}{\texttt{brian.mills@austin.utexas.edu}}
@@
 
-\hypertarget{data}{%
-\subsection{Data}\label{data}}
-
-We obtained multi-year baseball statistics from GREG??. Over the course
-of the sport's history, numerous statistics have been collected, and for
-our analysis, we focused on year-end statistics pertaining to the teams.
-To ensure consistency in our methodology, it was necessary for each team
-to have existed from 1900 until the end of the dataset in 2020. This
-timeframe of 120 years, denoted as T, encompassed a total of 16
-franchise teams, namely NYY, BOS, LAD, ATL, CHW, CHC, CIN, CLE, DET,
-BAL, SFG, OAK, PHI, PIT, STL, and MIN. The statistics of interest for
-these teams include runs, hits, home runs, balls, strikeouts, at-bats,
-stolen bases, number of games played in a season, attendance, runs
-against, hits against, home runs against, balls against, and strikeouts
-against. However, a few statistics, such as balls, stolen bases,
-strikeouts, and attendance, had missing data for certain years. To
-address this, we employed multiple imputations using classification and
-regression trees (cite mice/cart) to fill in the missing values and
-ensure consistent data across all analyses.
-
-\hypertarget{eras-and-dynasties}{%
-\subsection{Eras and Dynasties}\label{eras-and-dynasties}}
-
-\hypertarget{eras}{%
-\subsubsection{Eras}\label{eras}}
-
-Since we are interested in finding where in time these different eras of
-baseball have occurred, we first examine each baseball statistic of
-interest as their own analysis for all 16 teams. Those baseball
-statistics are home runs, strikeouts, balls, stolen bases, runs, and
-attendance which were all standardized to the number of games played in
-a season since we are interested in seeing how the different statistics
-change from season to season. Using the methodology from Cho and Fry
-(CITE) a thresholding parameter was determined from bootstrapping and
-then used to find where change points in mean and variance existed in
-the panel of teams data of a given baseball statistic. To examine the
-idea of changes throughout teams and types of statistics (initially),
-for each season the average baseball statistic was found from all the
-teams that year and then an average time series was created of each of
-the types of statistics of interest. Then the change point methodology
-was performed on those 6 time series of average statistic types.
-
-\#\#\#Dynasties
-
-Continuing our investigation, we explored the concept of ``dynasties''
-within modern baseball teams, irrespective of their length of a teams
-existence. A dynasty, in this context, refers to a period characterized
-by significant success or potentially hardships. Such a dynasty is
-identified when a team experiences a noticeable change across all their
-statistics in a given season, whether it be in a positive or negative
-direction. This approach aligns well with our methodology, as it
-involves examining panels of time series data encompassing all
-statistics for a singular team. The statistics of interest for this
-analysis include runs, hits, home runs, balls, strikeouts, runs against,
-hits against, home runs against, balls against, and strikeouts against.
-To ensure comparability, these statistics were standardized using the
-season average and standard deviation derived from all teams present in
-each respective season. By standardizing the statistics, we capture a
-team's offensive and defensive performance. Notably, if a team exhibits
-a high and positive offensive performance, it can be balanced to zero by
-large and negative defensive statistics when standardized. Therefore, if
-a team excels in both aspects of the game, it would have numerous time
-points above zero, whereas underperforming teams would display many
-below-zero values.For each team, we examined their ten statistics,
-analyzing both mean and variance changes to identify potential change
-points. By scrutinizing these indicators, we aimed to uncover shifts in
-a team's performance and assess their significance within the context of
-dynasties.
-
 \hypertarget{results}{%
@@
 
-\hypertarget{tables}{%
-\subsection{Tables}\label{tables}}
+\begin{verbatim}
+## [1]  20  46  67  94 106
+\end{verbatim}
 
-\begin{longtable}[]{@{}lll@{}}
-\toprule\noalign{}
-Mean/Variance CPT & Franchise & Change Point Year \\
-\midrule\noalign{}
+\begin{verbatim}
+##  b 
+## 48
+\end{verbatim}
+
+\begin{center}\includegraphics{paper_files/figure-latex/unnamed-chunk-1-1} \end{center}
+
+\hypertarget{results-1}{%
+\subsection{Results}\label{results-1}}
+
+\begin{longtable}[]{@{}lrr@{}}
+\toprule()
+franch\_id & threshold & years \\
+\midrule()
 \endhead
-\bottomrule\noalign{}
-\endlastfoot
-mean & ANA & 1977 \\
-mean & ATL & 1900, 1990 \\
-mean & BAL & 1959, 1999 \\
-mean & BOS & 1918, 1937 \\
-mean & CHC & 1891 \\
-mean & CHW & 1970 \\
-mean & CIN & 1918, 1952 \\
-mean & CLE & 1944, 1959, 1993 \\
-mean & DET & 1958, 1988 \\
-mean & FLA & 2011 \\
-mean & HOU & 1999 \\
-mean & KCR & 1997 \\
-mean & LAD & 1940, 1961 \\
-mean & MIL & 1977, 1996 \\
-mean & MIN & 1959 \\
-mean & PHI & 1917, 1949 \\
-mean & PIT & 1939 \\
-mean & SDP & 1977 \\
-mean & SEA & 1986, 1999 \\
-mean & SFG & 1903, 1973 \\
-mean & STL & 1937 \\
-mean & TBD & 2007 \\
-mean & TEX & 1973, 1990 \\
-mean & WSN & 2011 \\
-variance & ATL & 1887 \\
-variance & BOS & 2008 \\
-variance & CHW & 1969 \\
-variance & CLE & 1966, 1981, 1993 \\
-variance & DET & 1942 \\
-variance & HOU & 2010 \\
-variance & NYM & 1974 \\
-variance & PIT & 1902, 2001 \\
-variance & SDP & 1977 \\
-variance & SFG & 1918 \\
-\end{longtable}
-
-\begin{longtable}[]{@{}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2000}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.2889}}
-  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5111}}@{}}
-\toprule\noalign{}
-\begin{minipage}[b]{\linewidth}\raggedright
-Mean/Variance CPT
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Baseball Statistic
-\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
-Change Point Year
-\end{minipage} \\
-\midrule\noalign{}
-\endhead
-\bottomrule\noalign{}
-\endlastfoot
-mean & att\_per\_game & 1945, 1975, 1992 \\
-mean & bb\_per\_game & 1927 \\
-mean & hr\_per\_game & 1920, 1946, 1967, 1993 \\
-mean & r\_per\_game & 1919, 1940 \\
-mean & sb\_per\_game & 1919 \\
-mean & so\_per\_game & 1956, 1993, 2009 \\
-variance & att\_per\_game & 1944, 2008 \\
-variance & hr\_per\_game & 1927, 1944, 1995, 2007 \\
-variance & r\_per\_game & 1935 \\
-variance & sb\_per\_game & 1919, 1966 \\
-variance & so\_per\_game & 1990 \\
+NYY & 4.801818 & 1918 \\
+BOS & 4.427752 & 1917 \\
+BOS & 4.427752 & 1929 \\
+BOS & 4.427752 & 1948 \\
+BOS & 4.427752 & 1994 \\
+LAD & 4.776867 & 1919 \\
+LAD & 4.776867 & 1955 \\
+ATL & 5.393778 & 1918 \\
+ATL & 5.393778 & 1949 \\
+CHW & 4.326192 & 1918 \\
+CHC & 4.277577 & 1919 \\
+CHC & 4.277577 & 1940 \\
+CIN & 4.249476 & 1929 \\
+CIN & 4.249476 & 1952 \\
+CLE & 4.405339 & 1920 \\
+CLE & 4.405339 & 1939 \\
+DET & 3.926054 & 1927 \\
+DET & 3.926054 & 1940 \\
+BAL & 4.641706 & 1919 \\
+BAL & 4.641706 & 1942 \\
+SFG & 4.319634 & 1919 \\
+SFG & 4.319634 & 1937 \\
+OAK & 3.819294 & 1920 \\
+OAK & 3.819294 & 1936 \\
+OAK & 3.819294 & 1951 \\
+PHI & 4.123104 & 1918 \\
+PHI & 4.123104 & 1937 \\
+PIT & 4.175793 & 1920 \\
+PIT & 4.175793 & 1946 \\
+STL & 5.444239 & 1919 \\
+STL & 5.444239 & 1941 \\
+MIN & 4.498666 & 1919 \\
+MIN & 4.498666 & 1942 \\
+\bottomrule()
 \end{longtable}

### 2023-07-19 manuscript_JQAS/manuscript_JQAS.Rnw
JQAS format.

\documentclass[USenglish]{article}	
% for 2-column layout use \documentclass[USenglish,twocolumn]{article}

\usepackage[utf8]{inputenc}				%(only for the pdftex engine)
%\RequirePackage[no-math]{fontspec}[2017/03/31]%(only for the luatex or the xetex engine)
\usepackage[big,online]{dgruyter}	%values: small,big | online,print,work
\usepackage{lmodern} 
\usepackage{microtype}
\usepackage[numbers,square,sort&compress]{natbib}

% New theorem-like environments will be introduced by using the commands \theoremstyle and \newtheorem.
% Please note that the environments proof and definition are already defined within dgryuter.sty.
\theoremstyle{dgthm}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{assertion}{Assertion}
\newtheorem{result}{Result}
\newtheorem{conclusion}{Conclusion}

\theoremstyle{dgdef}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}
\newtheorem{remark}{Remark}

\begin{document}

%%%--------------------------------------------%%%
	\articletype{Research Article}
	\received{Month	DD, YYYY}
	\revised{Month	DD, YYYY}
  \accepted{Month	DD, YYYY}
  \journalname{De~Gruyter~Journal}
  \journalyear{YYYY}
  \journalvolume{XX}
  \journalissue{X}
  \startpage{1}
  \aop
  \DOI{10.1515/sample-YYYY-XXXX}
%%%--------------------------------------------%%%

\title{Insert your title here}
\runningtitle{Short title}
%\subtitle{Insert subtitle if needed}

\author*[1]{First Author}
%\ use * to mark the author as the corresponding author
\author[2]{Second Author}
\author[2]{Third Author} 
\runningauthor{F.~Author et al.}
\affil[1]{\protect\raggedright 
Institution, Department, City, Country of first author, e-mail: author\_one@xx.yz}
\affil[2]{\protect\raggedright 
Institution, Department, City, Country of second author and third author, e-mail: author\_two@xx.yz, author\_three@xx.yz}

%\communicated{...}
%\dedication{...}

\abstract{Please insert your abstract here (approx. 200 words). Remember that online systems rely heavily on the content of titles and abstracts to identify articles in electronic bibliographic databases and search engines. Abstracts give a clear indication of the nature and the range of the results in the paper.}

\keywords{Please insert your keywords here, separated by commas.}

\maketitle

\section{Introduction}

The first professional baseball team in the United States, the Cinicnnati Red Stockings, was formed in 1869 (@BBHOF1869).   Many leagues came and went in the late 1800s, but National League (NL), formed in 1876, emerged as the predominant league of the time.  A few decades later, the American League (AL) began growing in popularity and eventually reached an agreement with the NL to be the two major leagues of baseball with the winner of each league playing in the World Series starting in 1903.

Throughout the history of baseball in the United States, the game has gone through many changes and distinct eras.  For example, the time period between approximately 1900-1919 is often referred to as the "Dead Ball Era" and was marked by low scoring games and dominant pitching.  Another more recent example would be the "Steroid Era" which lasted from approximately 1994 through 2005 and was characterized by a rapid increase in power hitting largely attributed to players using performance enhancing drugs.

More specifically, @Woltring2018 mentions six eras of modern baseball": 
"Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras. A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005)."  @Woltring2018 notes that they name a seventh era after 2006, which they term the Post Steroid Era.

While many baseball writers have attempted to define the different eras of baseball, there has also been some academic work that has sought to empirically define eras in baseball.  @Groothius2017 looked for structural breaks in univariate times series of performance measures over the period from 1871-2020.  They analyzed four statistics: slugging percentage (SLG, home run (HR) rate, batting average (BA), and runs batted in (RBI) rate.   For each of these statistics, the computed the mean and standard deviation (SD) across all players who had at least 100 at bats in a given season to yield a univariate time series for each of these statistics.  They then used the Lagrange Multiplier (LM) unit root test proposed in @LeeandStrazicich2003 to find structural breaks.  They identified structural breaks in slugging percentage in 1921 and 1992, the first of which marks the end of the Dead Ball Era and the latter corresponding to the start of the steroid era.

@LeeFort2005 looks for structural changes in competitive balance of the two league American and National.  Use methods from Andrews1993, Bai1997, 1999 and Bai and Perron 1998 and 2003 to look for break points between 1901 -1999.  Theuy measure competitive balance in two ways: 1) @Noll1988 and @Scully1989 and 2) Lee 2004. They find break point sin competitive balance in 1912, 1926, and 1933 for the NL and in 1926 and 1957 in the AL.

Baseball is not the only sport where this type of analysis has been applied.  @PalaciosHuerta2004 looked for structural changes in soccer using data from British soccer leagues through 1996.  They had data from division I (Premier League) and II starting in the late 1800s, and data from lower divisions III and IV from just after WWII starting in 1947.  They identify a number of change points.  Notably they identify a change point in the mean of margin of victory in 1925 related to the change in the definition of offsides (changed from 3 players to 2 players), an change points in the variability of number of goals in the early 1980s and 1992 corresponding to the change in number of points for a win (i.e. went from 2 points to 3 points) and a change in the backpass rule, respectively.  @FortLee2007 looked for structural breaks in major North American sports other than baseball (i.e. basketball (NBA), hockey (NHL), and American football (NFL)).  They identified a number of change points related to competitive balance in each sport that often, though not always, correspond to league expansion, league mergers, or other major events in a sport (e.g. increased number of foreign players in the NBA in the late 1990s/early 2000s).

% # WHAT DID THEY FIND?   ADD MORE HERE.

All of the previous work mentioned here focuses on change point analysis in {\bf univariate} time series. However, recent methodolgical developments in change point analysis allow for change point analysis in {\bf multivariate} time series, which is the focus of this current manuscript.  This work leverages techniques such as the Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014) to look for change points in Major League Baseball at two main levels.  First, we look for structural changes in the league as a whole across teams to empirically define different eras in baseball.  Second, we look for change points within a team to determine eras of a team.  This second analysis can be used to identify the beginning and end of so called "dynasties", periods of sustained excellent performance by a team.  In addition, we can also identify the opposite, sustained periods of poor performance.

@LeeFort2008: Attendance and the Uncertainty-of-Outcome Hypothesis in Baseball.  Identify Break points in attendance in 1918 and 1945 for both leagues.  For AL only: 1963, 1978, 1994.  For NL only: 1976.  Using Bai and Perron From table 3 in Mills and Fort 2014
@MillsFort2014: LEAGUE-LEVEL ATTENDANCE AND OUTCOME UNCERTAINTY IN U.S. PRO SPORTS LEAGUES.  Looks at NHL, NFL, NBA.  Rottenburg 1956 looked at baseball.  Identify break points in NBA, NFL, and NHL.  Table 3
@MillsSalaga2015: NCAA Basketball: League balance using stats.  
@MillsFort2018: team-level: Attendence.  
@SalagaFort2017: College football

% <!-- Just Notes:  -->
% <!-- Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258 -->
% <!-- @PalaciosHuerta2004 Structural Breaks in soccer -->
% 
% 
% <!-- @Nieswiadomy2012 Was there a structural break in Barry Bonds’s Bat?  Looking for structural breaks in Bonds OPS times series. Use the uni root test on monthly data from 86-07.  The identify break points in June 1993 and Sept 2000.  -->
% 
% 
% <!-- In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014). -->
% 
% <!-- https://onlinelibrary.wiley.com/doi/epdf/10.1111/j.1465-7295.2007.00026.x -->
% <!-- Fort and Lee 2007: Looks for structural breaks in NBA, NHL, and NFL -->
% 
% 
% 
% <!-- https://www.baseball-reference.com/bullpen/History_of_baseball_in_the_United_States -->
% 
% <!-- https://thesportjournal.org/article/examining-perceptions-of-baseballs-eras/ -->

SOME MORE STUFF

% <!-- https://pubmed.ncbi.nlm.nih.gov/30747582/? -->
% 
% 
% <!-- The goal of this work is to use multivariate change point analysis to identify change points in MLB as a whole to objectively identify eras in the games history, but also to look for change points to identify team specific eras.   -->
% 
% <!-- The work that is the most closely related to our own is that of @Groothius2017, which looked for structural breaks (Note: the terms "structural breaks" and "change points" are used interchangeably in this manuscript) in univariate times series.   -->
% 
% <!-- univariate time series of the means and SD's of several statistics (i.e. SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrange Multiplier (LM) unit root test @LeeandStrazicich2003.  Notably they find structural breaks in slugging percentage in 1921 and 1992.  In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014). -->
% 
% 
% <!-- https://journals.sagepub.com/doi/10.1177/152700250100200204 -->
% 
% <!-- Deadball era: https://sabr.org/journal/article/the-rise-and-fall-of-the-deadball-era/ -->
% 
% <!-- http://mlb.mlb.com/mlb/history/mlb_history_people.jsp?story=com -->
% 
% <!-- https://www.nytimes.com/2011/09/25/sports/baseball/scoring-in-baseball-returns-to-dead-ball-levels.html?_r=1& -->
% 
% <!-- https://bleacherreport.com/articles/1676509-the-evolution-of-the-baseball-from-the-dead-ball-era-through-today -->
% 
% <!-- @FortLee2006:  -->
% <!-- Fort R., Lee Y. H. (2006). Stationarity and major league baseball attendance analysis. Journal of Sports Economics, 7, 408–415. -->
% 
% <!-- Fort R., Lee Y. H. (2007). Structural change, competitive balance, and the rest of the major leagues. Economic Inquiry, 45, 519–532. -->
% 
% <!-- Lee Y. H., Fort R. (2005). Structural change in baseball’s competitive balance: The great depression, team location, and racial integration. Economic Inquiry, 43, 158–169. -->
% 
% 
% <!-- @Nieswiadomy2012 Was there a structural break in Barry Bonds’s Bat?  Looking for structural breaks in Bonds OPS times series. Use the uni root test on monthly data from 86-07.  The identify break points in June 1993 and Sept 2000.   -->
% 
% <!-- degruyter.com/document/doi/10.1515/1559-0410.1305/html -->
% 
% <!-- Palacios-Huerta I. (2004). Structural changes during a Century of the World’s most popular sport. Statistical Methods and Applications, 13, 241–258 -->
% <!-- @PalaciosHuerta2004 Structural Breaks in soccer -->
% 
% <!-- @Groothius2017 is the most closely related work to ours.  They use univariate time series of the means and SD's of several statistics (i.e. SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrane Multiplier (LM) unit root test @LeeandStrazicich2003.  Notably they find structural breaks in slugging percentage in 1921 and 1992.  In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).   -->
% 
% <!-- Fort, Rodney and Young Hoon Lee (2006). “Stationarity and Major League Baseball Attendance Analysis.” Journal of Sports Economics, 7 (4), 408–415.  -->
% 
% <!-- Fort, Rodney and Young Hoon Lee (2007). “Structural Change, Competitive Balance, and the Rest of the Major Leagues.” Economic Inquiry, 45 (3), 519-532.  -->
% 
% <!-- Lee, Young Hoon, and Rodney Fort (2005). “Structural Change in Baseball’s Competitive Balance: The Great Depression, Team Location, and Racial Integration.” Economic Inquiry, 43, 158-169.  -->
% 
% <!-- Sommers, P. (2008). “The Changing Hitting Performance Profile in Major League Baseball, 1966-2006.” Journal of Sports Economics, 9, 435-440.  -->
% 
% 
% 
% 
% 
% 
% 
% <!-- @Berry1999bridging they talk about bridging eras.   -->
% 
% <!-- When did the steroids era start: https://www.espn.com/mlb/topics/_/page/the-steroids-era#:~:text=Unlike%20other%20MLB%20%22eras%2C%22,leaguewide%20PED%20testing%20until%202003. -->
% 
% <!-- Traditional wisdom: -->
% <!-- 1900-1919 dead ball era -->
% 
% <!-- From Woltring: -->
% <!-- "Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras. A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005) (17). This study runs through the 2011 season and a seventh era will be added and labeled the Post Steroid Era (2006-2011)" -->
% 
% 
% <!-- https://www.baseball-reference.com/bullpen/Deadball_Era -->
% <!-- 1901-1920 -->
% 
% <!-- Mound was lowered in december 1968.   -->
% <!-- https://www.baseball-reference.com/bullpen/Pitcher%27s_mound -->
% 
% <!-- https://www.espn.com/mlb/story/_/id/33238595/major-league-baseball-stops-testing-players-steroids-nearly-20-years-report-says -->
% 
% <!-- https://www.semanticscholar.org/paper/Was-There-a-Structural-Break-in-Barry-Bonds's-Bat-Nieswiadomy-Strazicich/0a32effa10a3d26ea4675efcca3f13229efe0f7d -->
%

\section{Methods}
@ChoFryzlewwicz2014 and @Cho2016

R Pacakge: @hdbinseg

A commonly employed method for detecting changes over time involves the CUSUM statistic in binary segmentation, which creates a tree diagram, dividing the time series after successfully identifying a change point until no further change points are detected. The magnitude of the CUSUM difference between two segments generally indicates a potential change point, depending on the assumptions and statistical tests applied to the time series. While this process works well for a single time series, detecting change points across a panel of multiple time series requires a more comprehensive approach. @Cho2016 introduced the double CUSUM (DC) algorithm, assuming there are $n$ time series, all of the same length $T$, with unknown common change point locations. This algorithm utilizes the CUSUM values of multiple time series $(j = 1, \dots, n)$ to cross-sectionally compare time segments for potential change points.

To achieve this, Cho's DC statistic involves ordered CUSUM values for a given time location, favoring larger values as potential change point candidates. It incorporates a partial thresholding of the smaller ordered CUSUM values, effectively representing a scaled average of the most prominent values at a given location across all "n" time series. Maximizing the DC statistic across all time series and potential time locations results in a test statistic for detecting a change point. This test statistic is then compared against a specific testing criterion, denoted as $\pi_{n,T}^\phi$, leading to the partitioning of time series and the formation of a tree structure based on the detected change points.

To determine the testing criterion $\pi_{n,T}^\phi$, a Generalized Dynamic Factor Model (GDFM) bootstrapping algorithm is employed, accounting for potential correlations within and between high-dimensional time series. This methodology enables the detection of second-order change points, utilizing methodology from @ChoFryzlewwicz2014 Haar wavelet periodograms and cross-periodograms, as opposed to the traditional CUSUM statistic. For further details, refer to @Cho2016 and @ChoFryzlewwicz2014.

A commonly employed method for detecting changes over time involves the
CUSUM statistic in binary segmentation, which creates a tree diagram,
dividing the time series after successfully identifying a change point
until no further change points are detected. The magnitude of the CUSUM
difference between two segments generally indicates a potential change
point, depending on the assumptions and statistical tests applied to the
time series. While this process works well for a single time series,
detecting change points across a panel of multiple time series requires
a more comprehensive approach. H. Cho \cite{Cho2016} introduced the double CUSUM (DC) algorithm, assuming there are $n$ time series, all of the same length
$T$, with unknown common change point locations. This algorithm
utilizes the CUSUM values of multiple time series $(j = 1, \dots, n)$
to cross-sectionally compare time segments for potential change points.

To achieve this, Cho's DC statistic involves ordered CUSUM values for a
given time location, favoring larger values as potential change point
candidates. It incorporates a partial thresholding of the smaller
ordered CUSUM values, effectively representing a scaled average of the
most prominent values at a given location across all ``n'' time series.
Maximizing the DC statistic across all time series and potential time
locations results in a test statistic for detecting a change point. This
test statistic is then compared against a specific testing criterion,
denoted as $\pi_{n,T}^\phi$, leading to the partitioning of time
series and the formation of a tree structure based on the detected
change points.

To determine the testing criterion $\pi_{n,T}^\phi$, a Generalized
Dynamic Factor Model (GDFM) bootstrapping algorithm is employed,
accounting for potential correlations within and between
high-dimensional time series. This methodology enables the detection of
second-order change points, utilizing methodology from H. Cho and
Fryzlewicz (\cite{ChoFryzlewwicz2014}). Haar
wavelet periodograms and cross-periodograms, as opposed to the
traditional CUSUM statistic. For further details, refer to H. Cho
(\cite{Cho2016}) and H. Cho and Fryzlewicz
(\cite{ChoFryzlewwicz2014}).

\section{Data}
We obtained multi-year baseball statistics from GREG??. Over the course of the sport's history, numerous statistics have been collected, and for our analysis, we focused on year-end statistics pertaining to the teams. To ensure consistency in our methodology, it was necessary for each team to have existed from 1900 until the end of the dataset in 2020. This timeframe of 120 years, denoted as T, encompassed a total of 16 franchise teams, namely NYY, BOS, LAD, ATL, CHW, CHC, CIN, CLE, DET, BAL, SFG, OAK, PHI, PIT, STL, and MIN. The statistics of interest for these teams include runs, hits, home runs, balls, strikeouts, at-bats, stolen bases, number of games played in a season, attendance, runs against, hits against, home runs against, balls against, and strikeouts against. However, a few statistics, such as balls, stolen bases, strikeouts, and attendance, had missing data for certain years. To address this, we employed multiple imputations using classification and regression trees (cite mice/cart) to fill in the missing values and ensure consistent data across all analyses.

\section{Methods}
<<echo = FALSE, message = FALSE, warning = FALSE>>=
library(tidyverse)
library(janitor)
library(hdbinseg)

teams <- read_csv("/Users/gregorymatthews/Dropbox/baseball_cpt/data/Teams.csv") %>% 
  clean_names()

team_names <- teams %>% 
  group_by(franch_id) %>% 
  count() %>% 
  filter(n > 120) %>% 
  pull(franch_id)

# teams %>% 
#   filter(franch_id %in% team_names) %>% 
#   group_by(franch_id) %>% 
#   summarise(min(year_id)) ## filter after 1901

old_teams_full <- teams %>% 
  filter(franch_id %in% team_names & year_id > 1900) %>% 
  select(year_id, franch_id, w, l, r, ab, h, hr, bb, so)

teams_wide <- old_teams_full %>% 
  pivot_wider(id_cols = franch_id, 
              names_from = year_id, 
              values_from = hr)

## find thresholding value for cpt using bootstrapping
thrs <- dcbs.thr(as.matrix(teams_wide[-1]), interval = c(1, dim(teams_wide[-1])[2]))
### if I lower the threshold more changepoints
hr_cpt_dcbs <- dcbs.alg(as.matrix(teams_wide[-1]), cp.type = 1, thr = thrs-1)
hr_cpt_dcbs$ecp
hr_cpt_sbs <- sbs.alg(as.matrix(teams_wide[-1]), cp.type = 1)         
hr_cpt_sbs$ecp

cpt_years <- as.numeric(names(teams_wide[-1])[hr_cpt_dcbs$ecp])

ggplot(old_teams_full, aes(year_id, hr, color = franch_id)) +
  geom_line() +
  geom_vline(xintercept = cpt_years)

@

\section{results}
%
% # <<>>=
% # load("./baseball_cpt/data/teams_cpt_time_df.RData")
% # teams_cpt_time_df
% # @
%

NYY 1918: Babe Ruth got to the Yankees in 1920.  And starting in 1919 they had exactly one losing seasons between 1919 and 1965.  
BOS 1917, 1929, 1948, 1994: 
Their last world series win in the 1900s was in 1918.  
I don't know 1929. 
1948 there was a big jump in runs?  
1994: Strike year.

LAD: 1919, 1955

ATL 1918 1949

CHW: 1919: Blaack Sox Scandal

CHC: 1919 1940
1940: War.

Cin: 1929 1952

CLE 1920 1939
Cleveland won the world series in 1920. Major change in offensive output.

DET 1927 1940

BAL 1919 1942

SFG 1919 1937

OAK 1920 1936 1951

PHI 1918 1937

PIT 1920 1946

STL 1919 1941

MIN 1919 1942

Pearl Harbor was 1941.  So US was in war in 1942.

\section{Acknowledgements}
We thank Michael Lopez for suggesting we do "something with change point analysis".

\section{Supp}

All code for reproducing the analyses in this paper is publicly available at https://github.com/menawhalen/baseball\_cpt

\begin{acknowledgement}
  Please insert acknowledgments of the assistance of colleagues or similar notes of appreciation here.
\end{acknowledgement}

\begin{funding}
  Please insert information concerning research grant support here (institution and grant number). Please provide for each funder the funder’s DOI according to https://doi.crossref.org/funderNames?mode=list.
\end{funding}

\bibliographystyle{rss}
\bibliography{references.bib}

\end{document}

### 2023-07-19 manuscript_JQAS/manuscript_JQAS.Rnw
JQAS format.

@@
 
-\author*[1]{First Author}
+\author*[1]{Mena Q.Whalen}
 %\ use * to mark the author as the corresponding author
-\author[2]{Second Author}
-\author[2]{Third Author} 
-\runningauthor{F.~Author et al.}
+\author[2]{Brian Mills}
+\author[3]{Gregory J. Matthews} 
+\runningauthor{Whalen et al.}
 \affil[1]{\protect\raggedright 
-Institution, Department, City, Country of first author, e-mail: author\_one@xx.yz}
+Loyola University Chicago, Department of Mathematics and Statistics, Chicago, IL, USA, e-mail: mwhalen@luc.edu}
 \affil[2]{\protect\raggedright 
-Institution, Department, City, Country of second author and third author, e-mail: author\_two@xx.yz, author\_three@xx.yz}
+Texas, Department, City, Country of second author and third author, e-mail: author\_two@xx.yz, author\_three@xx.yz}
+\affil[3]{\protect\raggedright 
+Loyola University Chicago, Department of Mathematics and Statistics, Chicago, IL, USA, e-mail: gmatthews1@luc.edu}
 
@@
 
-The first professional baseball team in the United States, the Cinicnnati Red Stockings, was formed in 1869 (@BBHOF1869).
+The first professional baseball team in the United States, the Cinicnnati Red Stockings, was formed in 1869 (\cite{BBHOF1869}).
 Many leagues came and went in the late 1800s, but National League (NL), formed in 1876, emerged as the predominant league of the time.
@@
 
-More specifically, @Woltring2018 mentions six eras of modern baseball": 
+More specifically, \cite{Woltring2018} mentions six eras of modern baseball": 
 "Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras.
-A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005)."  @Woltring2018 notes that they name a seventh era after 2006, which they term the Post Steroid Era.
+A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005)."  \cite{Woltring2018} notes that they name a seventh era after 2006, which they term the Post Steroid Era.
 
 While many baseball writers have attempted to define the different eras of baseball, there has also been some academic work that has sought to empirically define eras in baseball.
-@Groothius2017 looked for structural breaks in univariate times series of performance measures over the period from 1871-2020.
+\cite{Groothius2017} looked for structural breaks in univariate times series of performance measures over the period from 1871-2020.
 They analyzed four statistics: slugging percentage (SLG, home run (HR) rate, batting average (BA), and runs batted in (RBI) rate.
@@
 
-@LeeFort2005 looks for structural changes in competitive balance of the two league American and National.
+\cite{LeeFort2005} looks for structural changes in competitive balance of the two league American and National.
 Use methods from Andrews1993, Bai1997, 1999 and Bai and Perron 1998 and 2003 to look for break points between 1901 -1999.
-Theuy measure competitive balance in two ways: 1) @Noll1988 and @Scully1989 and 2) Lee 2004.
+Theuy measure competitive balance in two ways: 1) \cite{Noll1988} and \cite{Scully1989} and 2) Lee 2004.
 They find break point sin competitive balance in 1912, 1926, and 1933 for the NL and in 1926 and 1957 in the AL.
@@
 Baseball is not the only sport where this type of analysis has been applied.
-@PalaciosHuerta2004 looked for structural changes in soccer using data from British soccer leagues through 1996.
+\cite{PalaciosHuerta2004} looked for structural changes in soccer using data from British soccer leagues through 1996.
 They had data from division I (Premier League) and II starting in the late 1800s, and data from lower divisions III and IV from just after WWII starting in 1947.
@@
 went from 2 points to 3 points) and a change in the backpass rule, respectively.
-@FortLee2007 looked for structural breaks in major North American sports other than baseball (i.e.
+\cite{FortLee2007} looked for structural breaks in major North American sports other than baseball (i.e.
 basketball (NBA), hockey (NHL), and American football (NFL)).

### 2023-07-20 manuscript_JQAS/manuscript_JQAS.Rnw
stuff

@@
 \affil[2]{\protect\raggedright 
-Texas, Department, City, Country of second author and third author, e-mail: author\_two@xx.yz, author\_three@xx.yz}
+University of Texas at Austin, Department of Kinesiology and Health Education, Austin, TX, USA, e-mail: brian.mills@austin.utexas.edu}
 \affil[3]{\protect\raggedright 
@@
 
-\abstract{Please insert your abstract here (approx.
-200 words).
-Remember that online systems rely heavily on the content of titles and abstracts to identify articles in electronic bibliographic databases and search engines.
-Abstracts give a clear indication of the nature and the range of the results in the paper.}
+\abstract{}
 
-\keywords{Please insert your keywords here, separated by commas.}
+\keywords{Baseball, Change Point Analysis}
 
@@
 
-The first professional baseball team in the United States, the Cinicnnati Red Stockings, was formed in 1869 (\cite{BBHOF1869}).
+The first professional baseball team in the United States, the Cincinnati Red Stockings, was formed in 1869 (\cite{BBHOF1869}).
 Many leagues came and went in the late 1800s, but National League (NL), formed in 1876, emerged as the predominant league of the time.
@@
 For each of these statistics, the computed the mean and standard deviation (SD) across all players who had at least 100 at bats in a given season to yield a univariate time series for each of these statistics.
-They then used the Lagrange Multiplier (LM) unit root test proposed in @LeeandStrazicich2003 to find structural breaks.
+They then used the Lagrange Multiplier (LM) unit root test proposed in \cite{LeeandStrazicich2003} to find structural breaks.
 They identified structural breaks in slugging percentage in 1921 and 1992, the first of which marks the end of the Dead Ball Era and the latter corresponding to the start of the steroid era.
@@
 Use methods from Andrews1993, Bai1997, 1999 and Bai and Perron 1998 and 2003 to look for break points between 1901 -1999.
-Theuy measure competitive balance in two ways: 1) \cite{Noll1988} and \cite{Scully1989} and 2) Lee 2004.
+They measure competitive balance in two ways: 1) \cite{Noll1988} and \cite{Scully1989} and 2) Lee 2004.
 They find break point sin competitive balance in 1912, 1926, and 1933 for the NL and in 1926 and 1957 in the AL.
@@
 However, recent methodolgical developments in change point analysis allow for change point analysis in {\bf multivariate} time series, which is the focus of this current manuscript.
-This work leverages techniques such as the Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014) to look for change points in Major League Baseball at two main levels.
+This work leverages techniques such as the Double CUSUM Binary Segmentation algorithm (\cite{Cho2016}) and the Sparsified Binary Segmentation algorithm (\cite{ChoFryzlewwicz2014}) to look for change points in Major League Baseball at two main levels.
 First, we look for structural changes in the league as a whole across teams to empirically define different eras in baseball.
@@
 
-@LeeFort2008: Attendance and the Uncertainty-of-Outcome Hypothesis in Baseball.
+% @LeeFort2008: Attendance and the Uncertainty-of-Outcome Hypothesis in Baseball.
 Identify Break points in attendance in 1918 and 1945 for both leagues.
@@
 Using Bai and Perron From table 3 in Mills and Fort 2014
-@MillsFort2014: LEAGUE-LEVEL ATTENDANCE AND OUTCOME UNCERTAINTY IN U.S.
+% @MillsFort2014: LEAGUE-LEVEL ATTENDANCE AND OUTCOME UNCERTAINTY IN U.S.
 PRO SPORTS LEAGUES.
@@
 Table 3
-@MillsSalaga2015: NCAA Basketball: League balance using stats.  
-@MillsFort2018: team-level: Attendence.  
-@SalagaFort2017: College football
+% @MillsSalaga2015: NCAA Basketball: League balance using stats.  
+% @MillsFort2018: team-level: Attendence.  
+% @SalagaFort2017: College football
 
@@
 
-SOME MORE STUFF
+%SOME MORE STUFF
 
@@
 \section{Methods}
-@ChoFryzlewwicz2014 and @Cho2016
+%@ChoFryzlewwicz2014 and @Cho2016
 
-R Pacakge: @hdbinseg
+R Pacakge: \cite{hdbinseg}
 
-A commonly employed method for detecting changes over time involves the CUSUM statistic in binary segmentation, which creates a tree diagram, dividing the time series after successfully identifying a change point until no further change points are detected.
+A commonly employed method for detecting changes over time involves the CUSUM statistic in binary segmentation (CITE XX), which creates a tree diagram, dividing the time series after successfully identifying a change point until no further change points are detected.
 The magnitude of the CUSUM difference between two segments generally indicates a potential change point, depending on the assumptions and statistical tests applied to the time series.
 While this process works well for a single time series, detecting change points across a panel of multiple time series requires a more comprehensive approach.
-@Cho2016 introduced the double CUSUM (DC) algorithm, assuming there are $n$ time series, all of the same length $T$, with unknown common change point locations.
+\cite{Cho2016} introduced the double CUSUM (DC) algorithm, assuming there are $n$ time series, all of the same length $T$, with unknown common change point locations.
 This algorithm utilizes the CUSUM values of multiple time series $(j = 1, \dots, n)$ to cross-sectionally compare time segments for potential change points.
 
-To achieve this, Cho's DC statistic involves ordered CUSUM values for a given time location, favoring larger values as potential change point candidates.
-It incorporates a partial thresholding of the smaller ordered CUSUM values, effectively representing a scaled average of the most prominent values at a given location across all "n" time series.
+To achieve this, the DC statistic from \cite{Cho2016} involves ordered CUSUM values for a given time location, favoring larger values as potential change point candidates.
+It incorporates a partial thresholding of the smaller ordered CUSUM values, effectively representing a scaled average of the most prominent values at a given location across all $n$ (WHY WAS "n" in quotes here?) time series.
 Maximizing the DC statistic across all time series and potential time locations results in a test statistic for detecting a change point.
@@
 
-To determine the testing criterion $\pi_{n,T}^\phi$, a Generalized Dynamic Factor Model (GDFM) bootstrapping algorithm is employed, accounting for potential correlations within and between high-dimensional time series.
-This methodology enables the detection of second-order change points, utilizing methodology from @ChoFryzlewwicz2014 Haar wavelet periodograms and cross-periodograms, as opposed to the traditional CUSUM statistic.
-For further details, refer to @Cho2016 and @ChoFryzlewwicz2014.
+To determine the testing criterion $\pi_{n,T}^\phi$, a Generalized Dynamic Factor Model (GDFM) bootstrapping algorithm is employed (CITE XX), accounting for potential correlations within and between high-dimensional time series.
+This methodology enables the detection of second-order change points, utilizing methodology from \cite{ChoFryzlewwicz2014} Haar wavelet periodograms and cross-periodograms, as opposed to the traditional CUSUM statistic.
+For further details, refer to \cite{Cho2016} and \cite{ChoFryzlewwicz2014}.
+These methods were implemented using the \texttt{hdbinseg} \cite{hdbinseg} in the R programming lanugage (\cite{R2023language}).
 
-A commonly employed method for detecting changes over time involves the
-CUSUM statistic in binary segmentation, which creates a tree diagram,
-dividing the time series after successfully identifying a change point
-until no further change points are detected. The magnitude of the CUSUM
-difference between two segments generally indicates a potential change
-point, depending on the assumptions and statistical tests applied to the
-time series. While this process works well for a single time series,
-detecting change points across a panel of multiple time series requires
-a more comprehensive approach. H. Cho \cite{Cho2016} introduced the double CUSUM (DC) algorithm, assuming there are $n$ time series, all of the same length
-$T$, with unknown common change point locations. This algorithm
-utilizes the CUSUM values of multiple time series $(j = 1, \dots, n)$
-to cross-sectionally compare time segments for potential change points.
-
-To achieve this, Cho's DC statistic involves ordered CUSUM values for a
-given time location, favoring larger values as potential change point
-candidates. It incorporates a partial thresholding of the smaller
-ordered CUSUM values, effectively representing a scaled average of the
-most prominent values at a given location across all ``n'' time series.
-Maximizing the DC statistic across all time series and potential time
-locations results in a test statistic for detecting a change point. This
-test statistic is then compared against a specific testing criterion,
-denoted as $\pi_{n,T}^\phi$, leading to the partitioning of time
-series and the formation of a tree structure based on the detected
-change points.
-
-To determine the testing criterion $\pi_{n,T}^\phi$, a Generalized
-Dynamic Factor Model (GDFM) bootstrapping algorithm is employed,
-accounting for potential correlations within and between
-high-dimensional time series. This methodology enables the detection of
-second-order change points, utilizing methodology from H. Cho and
-Fryzlewicz (\cite{ChoFryzlewwicz2014}). Haar
-wavelet periodograms and cross-periodograms, as opposed to the
-traditional CUSUM statistic. For further details, refer to H. Cho
-(\cite{Cho2016}) and H. Cho and Fryzlewicz
-(\cite{ChoFryzlewwicz2014}).
+%hdbinseg: Change-Point Analysis of High-Dimensional Time Series via Binary Segmentation
 
 \section{Data}
-We obtained multi-year baseball statistics from GREG??.
+We obtained multi-year baseball statistics from the Lahman database (\cite{Friendly2021Lahman}).
 Over the course of the sport's history, numerous statistics have been collected, and for our analysis, we focused on year-end statistics pertaining to the teams.
-To ensure consistency in our methodology, it was necessary for each team to have existed from 1900 until the end of the dataset in 2020.
+To ensure consistency in our methodology, it was necessary for each team to have existed from 1900 until the end of the dataset in 2020 (IS THIS BECAUSE THE CHO2016 method requires the time series to be the same length?).
 This timeframe of 120 years, denoted as T, encompassed a total of 16 franchise teams, namely NYY, BOS, LAD, ATL, CHW, CHC, CIN, CLE, DET, BAL, SFG, OAK, PHI, PIT, STL, and MIN.
@@
 However, a few statistics, such as balls, stolen bases, strikeouts, and attendance, had missing data for certain years.
-To address this, we employed multiple imputations using classification and regression trees (cite mice/cart) to fill in the missing values and ensure consistent data across all analyses.
+To address this, we employed multiple imputations using classification and regression trees (\cite{Vanbuuren1999flexible}) to fill in the missing values and ensure consistent data across all analyses.
+(HOW DID WE DO THRE IMPUTATION?
+Was is single imputation?
+)
 
-\section{Methods}
+\section{Results}
+\subsection{Eras}
+Since we are interested in finding where in time these different eras of baseball have occurred, we first examine each baseball statistic of interest as their own analysis for all 16 teams.
+Those baseball statistics are home runs, strikeouts, balls, stolen bases, runs, and attendance which were all standardized to the number of games played in a season since we are interested in seeing how the different statistics change from season to season.
+Using the methodology from Cho and Fry (CITE) a thresholding parameter was determined from bootstrapping and then used to find where change points in mean and variance existed in the panel of teams data of a given baseball statistic.
+To examine the idea of changes throughout teams and types of statistics (initially), for each season the average baseball statistic was found from all the teams that year and then an average time series was created of each of the types of statistics of interest.
+Then the change point methodology was performed on those 6 time series of average statistic types.
+
+Continuing our investigation, we explored the concept of "dynasties" within modern baseball teams, irrespective of their length of a teams existence.
+A dynasty, in this context, refers to a period characterized by significant success or potentially hardships.
+Such a dynasty is identified when a team experiences a noticeable change across all their statistics in a given season, whether it be in a positive or negative direction.
+This approach aligns well with our methodology, as it involves examining panels of time series data encompassing all statistics for a singular team.
+The statistics of interest for this analysis include runs, hits, home runs, balls, strikeouts, runs against, hits against, home runs against, balls against, and strikeouts against.
+To ensure comparability, these statistics were standardized using the season average and standard deviation derived from all teams present in each respective season.
+By standardizing the statistics, we capture a team's offensive and defensive performance.
+Notably, if a team exhibits a high and positive offensive performance, it can be balanced to zero by large and negative defensive statistics when standardized.
+Therefore, if a team excels in both aspects of the game, it would have numerous time points above zero, whereas underperforming teams would display many below-zero values.For each team, we examined their ten statistics, analyzing both mean and variance changes to identify potential change points.
+By scrutinizing these indicators, we aimed to uncover shifts in a team's performance and assess their significance within the context of dynasties.
+
+\begin{table}
+\begin{tabular}{ccc}
+mean/ var & TM & year \\
+mean & ANA & 1977
+\end{tabular}
+\caption{F}
+\begin{table}
+
+|Mean/Variance CPT     |Franchise | Change Point Year |
+|:--------|:---------|:----------------|
+|mean     |ANA       |1977             |
+|mean     |ATL       |1900, 1990       |
+|mean     |BAL       |1959, 1999       |
+|mean     |BOS       |1918, 1937       |
+|mean     |CHC       |1891             |
+|mean     |CHW       |1970             |
+|mean     |CIN       |1918, 1952       |
+|mean     |CLE       |1944, 1959, 1993 |
+|mean     |DET       |1958, 1988       |
+|mean     |FLA       |2011             |
+|mean     |HOU       |1999             |
+|mean     |KCR       |1997             |
+|mean     |LAD       |1940, 1961       |
+|mean     |MIL       |1977, 1996       |
+|mean     |MIN       |1959             |
+|mean     |PHI       |1917, 1949       |
+|mean     |PIT       |1939             |
+|mean     |SDP       |1977             |
+|mean     |SEA       |1986, 1999       |
+|mean     |SFG       |1903, 1973       |
+|mean     |STL       |1937             |
+|mean     |TBD       |2007             |
+|mean     |TEX       |1973, 1990       |
+|mean     |WSN       |2011             |
+|variance |ATL       |1887             |
+|variance |BOS       |2008             |
+|variance |CHW       |1969             |
+|variance |CLE       |1966, 1981, 1993 |
+|variance |DET       |1942             |
+|variance |HOU       |2010             |
+|variance |NYM       |1974             |
+|variance |PIT       |1902, 2001       |
+|variance |SDP       |1977             |
+|variance |SFG       |1918             |
+
+|Mean/Variance CPT   |Baseball Statistic     | Change Point Year                    |
+|:--------|:------------|:----------------------|
+|mean     |att_per_game |1945, 1975, 1992       |
+|mean     |bb_per_game  |1927                   |
+|mean     |hr_per_game  |1920, 1946, 1967, 1993 |
+|mean     |r_per_game   |1919, 1940             |
+|mean     |sb_per_game  |1919                   |
+|mean     |so_per_game  |1956, 1993, 2009       |
+|variance |att_per_game |1944, 2008             |
+|variance |hr_per_game  |1927, 1944, 1995, 2007 |
+|variance |r_per_game   |1935                   |
+|variance |sb_per_game  |1919, 1966             |
+|variance |so_per_game  |1990                   |
+
 <<echo = FALSE, message = FALSE, warning = FALSE>>=
@@
 
-\section{results}
 %
@@
 
+\section{Conclusions}
+
 \section{Acknowledgements}

### 2023-07-20 cpts_all_variables.R

@@ -156,3 +156,10 @@ cpt_szn <- bind_rows(tibble(year = cpt_years, cpt_stat = "mean"),
 ggplot(szn_ave) +
   geom_line(aes(year_id, value, color = stat)) +
   geom_vline(data = cpt_szn, aes(xintercept = year, linetype = cpt_stat))
+
+
+
+bind_rows(time_cpts, time_cpts_var) %>% 
+  group_by(change, variable) %>% 
+  summarise(cpt = paste(years, collapse = ", ")) %>% 
+  kableExtra::kable(format = "markdown")

### 2023-07-20 current_teams_cpts.R

@@ -109,3 +109,10 @@ std_teams %>%
   geom_line() +
   geom_vline(data = total_values, aes(xintercept = years, linetype = type)) +
   facet_wrap(~franch_id, scales = "free_x")
+
+
+## print out table of results
+total_values %>% 
+  group_by(type, franch_id) %>% 
+  summarise(cpt = paste(years, collapse = ", ")) %>% 
+  kableExtra::kable(format = "markdown")

### 2023-07-20 manuscript/paper.Rmd
I screwed up and fixed it.
JQAS manuscript.

--- 2023-07-19 manuscript/paper.Rmd
@@
 
-## Data
+A commonly employed method for detecting changes over time involves the
+CUSUM statistic in binary segmentation, which creates a tree diagram,
+dividing the time series after successfully identifying a change point
+until no further change points are detected. The magnitude of the CUSUM
+difference between two segments generally indicates a potential change
+point, depending on the assumptions and statistical tests applied to the
+time series. While this process works well for a single time series,
+detecting change points across a panel of multiple time series requires
+a more comprehensive approach. H. Cho
+(\protect\hyperlink{ref-Cho2016}{2016}) introduced the double CUSUM (DC)
+algorithm, assuming there are \(n\) time series, all of the same length
+\(T\), with unknown common change point locations. This algorithm
+utilizes the CUSUM values of multiple time series \((j = 1, \dots, n)\)
+to cross-sectionally compare time segments for potential change points.
 
+To achieve this, Cho's DC statistic involves ordered CUSUM values for a
+given time location, favoring larger values as potential change point
+candidates. It incorporates a partial thresholding of the smaller
+ordered CUSUM values, effectively representing a scaled average of the
+most prominent values at a given location across all ``n'' time series.
+Maximizing the DC statistic across all time series and potential time
+locations results in a test statistic for detecting a change point. This
+test statistic is then compared against a specific testing criterion,
+denoted as \(\pi_{n,T}^\phi\), leading to the partitioning of time
+series and the formation of a tree structure based on the detected
+change points.
+
+To determine the testing criterion \(\pi_{n,T}^\phi\), a Generalized
+Dynamic Factor Model (GDFM) bootstrapping algorithm is employed,
+accounting for potential correlations within and between
+high-dimensional time series. This methodology enables the detection of
+second-order change points, utilizing methodology from H. Cho and
+Fryzlewicz (\protect\hyperlink{ref-ChoFryzlewwicz2014}{2014}) Haar
+wavelet periodograms and cross-periodograms, as opposed to the
+traditional CUSUM statistic. For further details, refer to H. Cho
+(\protect\hyperlink{ref-Cho2016}{2016}) and H. Cho and Fryzlewicz
+(\protect\hyperlink{ref-ChoFryzlewwicz2014}{2014}).
+
+#Data
 We obtained multi-year baseball statistics from GREG??.
@@
 
-## Eras and Dynasties
+# Results
+```{r echo = FALSE}
+library(tidyverse)
+library(janitor)
+library(hdbinseg)
 
-### Eras
-Since we are interested in finding where in time these different eras of baseball have occurred, we first examine each baseball statistic of interest as their own analysis for all 16 teams.
-Those baseball statistics are home runs, strikeouts, balls, stolen bases, runs, and attendance which were all standardized to the number of games played in a season since we are interested in seeing how the different statistics change from season to season.
-Using the methodology from Cho and Fry (CITE) a thresholding parameter was determined from bootstrapping and then used to find where change points in mean and variance existed in the panel of teams data of a given baseball statistic.
-To examine the idea of changes throughout teams and types of statistics (initially), for each season the average baseball statistic was found from all the teams that year and then an average time series was created of each of the types of statistics of interest.
-Then the change point methodology was performed on those 6 time series of average statistic types.
+teams <- read_csv("/Users/gregorymatthews/Dropbox/baseball_cpt/data/Teams.csv") %>% 
+  clean_names()
 
-###Dynasties
+team_names <- teams %>% 
+  group_by(franch_id) %>% 
+  count() %>% 
+  filter(n > 120) %>% 
+  pull(franch_id)
 
-Continuing our investigation, we explored the concept of "dynasties" within modern baseball teams, irrespective of their length of a teams existence.
-A dynasty, in this context, refers to a period characterized by significant success or potentially hardships.
-Such a dynasty is identified when a team experiences a noticeable change across all their statistics in a given season, whether it be in a positive or negative direction.
-This approach aligns well with our methodology, as it involves examining panels of time series data encompassing all statistics for a singular team.
-The statistics of interest for this analysis include runs, hits, home runs, balls, strikeouts, runs against, hits against, home runs against, balls against, and strikeouts against.
-To ensure comparability, these statistics were standardized using the season average and standard deviation derived from all teams present in each respective season.
-By standardizing the statistics, we capture a team's offensive and defensive performance.
-Notably, if a team exhibits a high and positive offensive performance, it can be balanced to zero by large and negative defensive statistics when standardized.
-Therefore, if a team excels in both aspects of the game, it would have numerous time points above zero, whereas underperforming teams would display many below-zero values.For each team, we examined their ten statistics, analyzing both mean and variance changes to identify potential change points.
-By scrutinizing these indicators, we aimed to uncover shifts in a team's performance and assess their significance within the context of dynasties.
+# teams %>% 
+#   filter(franch_id %in% team_names) %>% 
+#   group_by(franch_id) %>% 
+#   summarise(min(year_id)) ## filter after 1901
 
-# Results
+old_teams_full <- teams %>% 
+  filter(franch_id %in% team_names & year_id > 1900) %>% 
+  select(year_id, franch_id, w, l, r, ab, h, hr, bb, so)
+
+teams_wide <- old_teams_full %>% 
+  pivot_wider(id_cols = franch_id, 
+              names_from = year_id, 
+              values_from = hr)
+
+## find thresholding value for cpt using bootstrapping
+thrs <- dcbs.thr(as.matrix(teams_wide[-1]), interval = c(1, dim(teams_wide[-1])[2]))
+### if I lower the threshold more changepoints
+hr_cpt_dcbs <- dcbs.alg(as.matrix(teams_wide[-1]), cp.type = 1, thr = thrs-1)
+hr_cpt_dcbs$ecp
+hr_cpt_sbs <- sbs.alg(as.matrix(teams_wide[-1]), cp.type = 1)         
+hr_cpt_sbs$ecp
+
+cpt_years <- as.numeric(names(teams_wide[-1])[hr_cpt_dcbs$ecp])
+
+ggplot(old_teams_full, aes(year_id, hr, color = franch_id)) +
+  geom_line() +
+  geom_vline(xintercept = cpt_years)
+
+```
+
+## Results
+```{r}
+load("./baseball_cpt/data/teams_cpt_time_df.RData")
+teams_cpt_time_df
+```
 

### 2023-07-20 manuscript_JQAS/manuscript_JQAS.Rnw
JQAS manuscript.

@@
 \begin{table}
-\begin{tabular}{ccc}
-mean/ var & TM & year \\
-mean & ANA & 1977
+\begin{tabular}{|r|r|}
+Franchise & Change Point Year(s) \\
+\hline
+ANA       & 1977             \\
+ATL       & 1900, 1990       \\
+BAL       & 1959, 1999      \\
+BOS       & 1918, 1937       \\
+CHC       & 1891             \\
+CHW       & 1970             \\
+CIN       & 1918, 1952       \\
+CLE       & 1944, 1959, 1993 \\
+DET       & 1958, 1988       \\
+FLA       & 2011             \\
+HOU       & 1999             \\
+KCR       & 1997             \\
+LAD       & 1940, 1961       \\
+MIL       & 1977, 1996       \\
+MIN       & 1959             \\
+PHI       & 1917, 1949       \\
+PIT       & 1939             \\
+SDP       & 1977             \\
+SEA       & 1986, 1999       \\
+SFG       & 1903, 1973       \\
+STL       & 1937             \\
+TBD       & 2007             \\
+TEX       & 1973, 1990       \\
+WSN       & 2011             \\
 \end{tabular}
-\caption{F}
+\caption{Mean change points for teams}
+\label{tab:mean:team}
+\end{table}
+
 \begin{table}
+\begin{tabular}{|r|r|}
+Franchise & Change Point Year(s) \\
+\hline
+ATL       &1887             \\
+BOS       &2008             \\
+CHW       &1969             \\
+CLE       &1966, 1981, 1993 \\
+DET       &1942             \\
+HOU       &2010             \\
+NYM       &1974             \\
+PIT       &1902, 2001       \\
+SDP       &1977             \\
+SFG       &1918             \\
+\end{tabular}
+\caption{Variance change points for teams}
+\label{tab:var:team}
+\end{table}
 
-|Mean/Variance CPT     |Franchise | Change Point Year |
-|:--------|:---------|:----------------|
-|mean     |ANA       |1977             |
-|mean     |ATL       |1900, 1990       |
-|mean     |BAL       |1959, 1999       |
-|mean     |BOS       |1918, 1937       |
-|mean     |CHC       |1891             |
-|mean     |CHW       |1970             |
-|mean     |CIN       |1918, 1952       |
-|mean     |CLE       |1944, 1959, 1993 |
-|mean     |DET       |1958, 1988       |
-|mean     |FLA       |2011             |
-|mean     |HOU       |1999             |
-|mean     |KCR       |1997             |
-|mean     |LAD       |1940, 1961       |
-|mean     |MIL       |1977, 1996       |
-|mean     |MIN       |1959             |
-|mean     |PHI       |1917, 1949       |
-|mean     |PIT       |1939             |
-|mean     |SDP       |1977             |
-|mean     |SEA       |1986, 1999       |
-|mean     |SFG       |1903, 1973       |
-|mean     |STL       |1937             |
-|mean     |TBD       |2007             |
-|mean     |TEX       |1973, 1990       |
-|mean     |WSN       |2011             |
-|variance |ATL       |1887             |
-|variance |BOS       |2008             |
-|variance |CHW       |1969             |
-|variance |CLE       |1966, 1981, 1993 |
-|variance |DET       |1942             |
-|variance |HOU       |2010             |
-|variance |NYM       |1974             |
-|variance |PIT       |1902, 2001       |
-|variance |SDP       |1977             |
-|variance |SFG       |1918             |
+Table \ref{tab:mean:team} says some shit.  
+Table \ref{tab:var:team} says some shit.
 
-|Mean/Variance CPT   |Baseball Statistic     | Change Point Year                    |
-|:--------|:------------|:----------------------|
-|mean     |att_per_game |1945, 1975, 1992       |
-|mean     |bb_per_game  |1927                   |
-|mean     |hr_per_game  |1920, 1946, 1967, 1993 |
-|mean     |r_per_game   |1919, 1940             |
-|mean     |sb_per_game  |1919                   |
-|mean     |so_per_game  |1956, 1993, 2009       |
-|variance |att_per_game |1944, 2008             |
-|variance |hr_per_game  |1927, 1944, 1995, 2007 |
-|variance |r_per_game   |1935                   |
-|variance |sb_per_game  |1919, 1966             |
-|variance |so_per_game  |1990                   |
+\begin{table}
+\begin{tabular}{|r|r|}
+Statistic & Change Point Year(s) \\
+\hline
+attpergame &1945, 1975, 1992       \\
+bbpergame  &1927                   \\
+hrpergame  &1920, 1946, 1967, 1993 \\
+rpergame   &1919, 1940             \\
+sbpergame  &1919                   \\
+sopergame  &1956, 1993, 2009       \\
+\end{tabular}
+\caption{Mean Change points for statistics}
+\label{tab:mean:stat}
+\end{table}
+
+\begin{table}
+\begin{tabular}{|r|r|}
+Statistic & Change Point Year(s) \\
+\hline
+attpergame &1944, 2008             \\
+hrpergame  &1927, 1944, 1995, 2007 \\
+rpergame   &1935                   \\
+sbpergame  &1919, 1966             \\
+sopergame  &1990                   \\
+\end{tabular}
+\caption{Variance Change points for statistics}
+\label{tab:var:stat}
+\end{table}
 

### 2023-07-20 manuscript_JQAS/manuscript_JQAS.Rnw
Some changes

@@
 However, recent methodolgical developments in change point analysis allow for change point analysis in {\bf multivariate} time series, which is the focus of this current manuscript.
-This work leverages techniques such as the Double CUSUM Binary Segmentation algorithm (\cite{Cho2016}) and the Sparsified Binary Segmentation algorithm (\cite{ChoFryzlewwicz2014}) to look for change points in Major League Baseball at two main levels.
+This work uses the Double CUSUM Binary Segmentation algorithm (\cite{Cho2016,ChoFryzlewicz2014}) to look for change points in Major League Baseball at two main levels.
 First, we look for structural changes in the league as a whole across teams to empirically define different eras in baseball.
@@
 % 
-% <!-- In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014). -->
+% <!-- In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewicz2014). -->
 % 
@@
 Notably they find structural breaks in slugging percentage in 1921 and 1992.
-In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).
+In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewicz2014).
 -->
@@
 
+\section{Data}
+%EXPLAIN DATA NEEDS FOR 1) ERAS and 2) DYNASTIES.
+We obtained multi-year baseball statistics from the Lahman database (\cite{Friendly2021Lahman}).
+Over the course of the sport's history, numerous statistics have been collected, and for our analysis, we focused on year-end statistics pertaining to the teams.
+In order to implement the DC algorithm, each time series is required to be the same length.
+As a result, when studying eras in our analysis we focus on a set of 16 teams that have all existed over the period from 1900 to 2020 because this period of time covers most of the history of organized professional baseball and there are enough teams that have all existed over this time period to perform meaningful analysis.
+If we had instead chosen to extend this period to, for instance, 1890 to 2020, we would only have had XX teams available.
+This timeframe of $T=120$ years encompassed a total of 16 franchise teams listed in table \ref{tab:120year:franch}.
+The statistics of interest for these teams include runs, hits, home runs, balls, strikeouts, at-bats, stolen bases, number of games played in a season, attendance, runs against, hits against, home runs against, balls against, and strikeouts against.
+However, a few statistics, such as balls, stolen bases, strikeouts, and attendance, had missing data for certain years.
+To address this, we employed multiple imputations using classification and regression trees (\cite{Vanbuuren1999flexible}) to fill in the missing values and ensure consistent data across all analyses.
+(HOW DID WE DO THRE IMPUTATION?
+Was is single imputation?
+)
+
+For the dynasty analysis we used teams over different periods of time for every year that each team has existed....MORE DETAILS.
+
+\begin{table}
+\begin{tabular}{|r|}
+Franchise  \\
+\hline
+NYY Full names 
+ NYY, BOS, LAD, ATL, CHW, CHC, CIN, CLE, DET, BAL, SFG, OAK, PHI, PIT, STL, and MIN\\
+\hline
+\end{tabular}
+\caption{}
+\label{tab:120year:franch}
+\end{table}
+
 \section{Methods}
-%@ChoFryzlewwicz2014 and @Cho2016
+%@ChoFryzlewicz2014 and @Cho2016
 
@@
 To determine the testing criterion $\pi_{n,T}^\phi$, a Generalized Dynamic Factor Model (GDFM) bootstrapping algorithm is employed (CITE XX), accounting for potential correlations within and between high-dimensional time series.
-This methodology enables the detection of second-order change points, utilizing methodology from \cite{ChoFryzlewwicz2014} Haar wavelet periodograms and cross-periodograms, as opposed to the traditional CUSUM statistic.
-For further details, refer to \cite{Cho2016} and \cite{ChoFryzlewwicz2014}.
-These methods were implemented using the \texttt{hdbinseg} \cite{hdbinseg} in the R programming lanugage (\cite{R2023language}).
+This methodology enables the detection of second-order change points, utilizing methodology from \cite{ChoFryzlewicz2014} Haar wavelet periodograms and cross-periodograms, as opposed to the traditional CUSUM statistic.
+For further details, refer to \cite{Cho2016} and \cite{ChoFryzlewicz2014}.
+These methods were implemented using the \texttt{hdbinseg} \cite{hdbinseg} in the R programming language (\cite{R2023language}).
 
-%hdbinseg: Change-Point Analysis of High-Dimensional Time Series via Binary Segmentation
+Discuss somewhere how we can find mean change points and ALSO variance change points.
 
-\section{Data}
-We obtained multi-year baseball statistics from the Lahman database (\cite{Friendly2021Lahman}).
-Over the course of the sport's history, numerous statistics have been collected, and for our analysis, we focused on year-end statistics pertaining to the teams.
-To ensure consistency in our methodology, it was necessary for each team to have existed from 1900 until the end of the dataset in 2020 (IS THIS BECAUSE THE CHO2016 method requires the time series to be the same length?).
-This timeframe of 120 years, denoted as T, encompassed a total of 16 franchise teams, namely NYY, BOS, LAD, ATL, CHW, CHC, CIN, CLE, DET, BAL, SFG, OAK, PHI, PIT, STL, and MIN.
-The statistics of interest for these teams include runs, hits, home runs, balls, strikeouts, at-bats, stolen bases, number of games played in a season, attendance, runs against, hits against, home runs against, balls against, and strikeouts against.
-However, a few statistics, such as balls, stolen bases, strikeouts, and attendance, had missing data for certain years.
-To address this, we employed multiple imputations using classification and regression trees (\cite{Vanbuuren1999flexible}) to fill in the missing values and ensure consistent data across all analyses.
-(HOW DID WE DO THRE IMPUTATION?
-Was is single imputation?
-)
+SEGUE parageraph Using eras and dynasties:::::::::
 
-\section{Results}
-\subsection{Eras}
 Since we are interested in finding where in time these different eras of baseball have occurred, we first examine each baseball statistic of interest as their own analysis for all 16 teams.
 Those baseball statistics are home runs, strikeouts, balls, stolen bases, runs, and attendance which were all standardized to the number of games played in a season since we are interested in seeing how the different statistics change from season to season.
-Using the methodology from Cho and Fry (CITE) a thresholding parameter was determined from bootstrapping and then used to find where change points in mean and variance existed in the panel of teams data of a given baseball statistic.
+Using the methodology from \cite{ChoFryzlewicz2014} a thresholding parameter was determined from bootstrapping and then used to find where change points in mean and variance existed in the panel of teams data of a given baseball statistic.
 To examine the idea of changes throughout teams and types of statistics (initially), for each season the average baseball statistic was found from all the teams that year and then an average time series was created of each of the types of statistics of interest.
@@
 By scrutinizing these indicators, we aimed to uncover shifts in a team's performance and assess their significance within the context of dynasties.
+
+%hdbinseg: Change-Point Analysis of High-Dimensional Time Series via Binary Segmentation
+
+\section{Results}
+\subsection{Eras}
+Blah blah 
+\subsection{Dynasties}
+Blah blah
 
@@
 \section{Conclusions}
+Baseball researchers
 

### 2023-07-26 manuscript_JQAS/manuscript_JQAS.tex
updating writings, citations, and coding

--- 2023-07-20 manuscript_JQAS/manuscript_JQAS.Rnw
@@
 %\RequirePackage[no-math]{fontspec}[2017/03/31]%(only for the luatex or the xetex engine)
-\usepackage[big,online]{dgruyter}	%values: small,big | online,print,work
+
 \usepackage{lmodern} 
@@
 \usepackage[numbers,square,sort&compress]{natbib}
+\usepackage[big,online]{dgruyter}	%values: small,big | online,print,work
 
@@
 
+\usepackage{Sweave}
 \begin{document}
+\input{manuscript_JQAS-concordance}
 
@@
 
-\author*[1]{Mena Q.Whalen}
+\author*[1]{Mena C.R. Whalen}
 %\ use * to mark the author as the corresponding author
@@
 \affil[1]{\protect\raggedright 
-Loyola University Chicago, Department of Mathematics and Statistics, Chicago, IL, USA, e-mail: mwhalen@luc.edu}
+Loyola University Chicago, Department of Mathematics and Statistics, Chicago, IL, USA, e-mail: mwhalen3@luc.edu}
 \affil[2]{\protect\raggedright 
@@
 
-<<echo = FALSE, message = FALSE, warning = FALSE>>=
-library(tidyverse)
-library(janitor)
-library(hdbinseg)
-
-teams <- read_csv("/Users/gregorymatthews/Dropbox/baseball_cpt/data/Teams.csv") %>% 
-  clean_names()
-
-team_names <- teams %>% 
-  group_by(franch_id) %>% 
-  count() %>% 
-  filter(n > 120) %>% 
-  pull(franch_id)
-
-# teams %>% 
-#   filter(franch_id %in% team_names) %>% 
-#   group_by(franch_id) %>% 
-#   summarise(min(year_id)) ## filter after 1901
-
-old_teams_full <- teams %>% 
-  filter(franch_id %in% team_names & year_id > 1900) %>% 
-  select(year_id, franch_id, w, l, r, ab, h, hr, bb, so)
-
-teams_wide <- old_teams_full %>% 
-  pivot_wider(id_cols = franch_id, 
-              names_from = year_id, 
-              values_from = hr)
-
-## find thresholding value for cpt using bootstrapping
-thrs <- dcbs.thr(as.matrix(teams_wide[-1]), interval = c(1, dim(teams_wide[-1])[2]))
-### if I lower the threshold more changepoints
-hr_cpt_dcbs <- dcbs.alg(as.matrix(teams_wide[-1]), cp.type = 1, thr = thrs-1)
-hr_cpt_dcbs$ecp
-hr_cpt_sbs <- sbs.alg(as.matrix(teams_wide[-1]), cp.type = 1)         
-hr_cpt_sbs$ecp
-
-cpt_years <- as.numeric(names(teams_wide[-1])[hr_cpt_dcbs$ecp])
-
-ggplot(old_teams_full, aes(year_id, hr, color = franch_id)) +
-  geom_line() +
-  geom_vline(xintercept = cpt_years)
-
-@
-
-%
-% # <<>>=
-% # load("./baseball_cpt/data/teams_cpt_time_df.RData")
-% # teams_cpt_time_df
-% # @
-%
+\begin{Schunk}
+\begin{Soutput}
+[1] "g"          "r"          "ab"         "sb"         "hr"        
+[6] "bb"         "so"         "attendance"
+\end{Soutput}
+\begin{Soutput}
+   year_id  franch_id          g          r         ab         sb         hr 
+         0          0          0          0          0          0          0 
+        bb         so attendance 
+         0         16          0 
+\end{Soutput}
+\begin{Soutput}
+ iter imp variable
+  1   1  bb  so  sb  attendance
+  1   2  bb  so  sb  attendance
+  1   3  bb  so  sb  attendance
+  1   4  bb  so  sb  attendance
+  1   5  bb  so  sb  attendance
+  2   1  bb  so  sb  attendance
+  2   2  bb  so  sb  attendance
+  2   3  bb  so  sb  attendance
+  2   4  bb  so  sb  attendance
+  2   5  bb  so  sb  attendance
+  3   1  bb  so  sb  attendance
+  3   2  bb  so  sb  attendance
+  3   3  bb  so  sb  attendance
+  3   4  bb  so  sb  attendance
+  3   5  bb  so  sb  attendance
+  4   1  bb  so  sb  attendance
+  4   2  bb  so  sb  attendance
+  4   3  bb  so  sb  attendance
+  4   4  bb  so  sb  attendance
+  4   5  bb  so  sb  attendance
+  5   1  bb  so  sb  attendance
+  5   2  bb  so  sb  attendance
+  5   3  bb  so  sb  attendance
+  5   4  bb  so  sb  attendance
+  5   5  bb  so  sb  attendance
+\end{Soutput}
+\end{Schunk}
 

### 2023-08-17 manuscript_JQAS/manuscript_JQAS.Rnw
Adding Brians update

@@
 \begin{document}
+%\SweaveOpts{concordance=TRUE}
 
@@
 For example, the time period between approximately 1900-1919 is often referred to as the "Dead Ball Era" and was marked by low scoring games and dominant pitching.
-Another more recent example would be the "Steroid Era" which lasted from approximately 1994 through 2005 and was characterized by a rapid increase in power hitting largely attributed to players using performance enhancing drugs.
-
-More specifically, \cite{Woltring2018} mentions six eras of modern baseball": 
-"Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras.
+Another more recent example would be the "Steroid Era" which lasted from approximately 1994 through 2005 and was characterized by a rapid increase in power hitting often attributed to players using performance enhancing drugs.
+More specifically, \cite{Woltring2018} mentions six eras of modern baseball: "Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras.
 A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005)."  \cite{Woltring2018} notes that they name a seventh era after 2006, which they term the Post Steroid Era.
 
-While many baseball writers have attempted to define the different eras of baseball, there has also been some academic work that has sought to empirically define eras in baseball.
-\cite{Groothius2017} looked for structural breaks in univariate times series of performance measures over the period from 1871-2020.
+Although the determination of eras tends to be inexact and arbitrary, identification of regimes between large changes can be helpful to understand innovations in gameplay, management, and athleticism.
+For this reason, academic work has sought to empirically identify structural changes in the skills and gameplay, competitive balance, and fan demand (attendance).
+Specifically, \cite{Groothius2017} "let the data speak", looking for structural breaks in univariate times series of performance measures over the period from 1871-2020.
 They analyzed four statistics: slugging percentage (SLG, home run (HR) rate, batting average (BA), and runs batted in (RBI) rate.
@@
 They identified structural breaks in slugging percentage in 1921 and 1992, the first of which marks the end of the Dead Ball Era and the latter corresponding to the start of the steroid era.
-
-\cite{LeeFort2005} looks for structural changes in competitive balance of the two league American and National.
-Use methods from Andrews1993, Bai1997, 1999 and Bai and Perron 1998 and 2003 to look for break points between 1901 -1999.
-They measure competitive balance in two ways: 1) \cite{Noll1988} and \cite{Scully1989} and 2) Lee 2004.
-They find break point sin competitive balance in 1912, 1926, and 1933 for the NL and in 1926 and 1957 in the AL.
+\cite{LeeFort2005} look for structural changes in competitive balance of the two leagues:  American and National using methods from (Bai and Perron 1998 and 2003) and others (\cite{Andrews1993,bai1997, bai1999}) to identify break points between 1901-1999.
+They measure competitive balance using the classical Noll_scully ratio (\cite{Noll1988,Scully1989} also adjusting for ties (Lee 2004).
+They find break points in competitive balance in 1912, 1926, and 1933 for the NL and in 1926 and 1957 in the AL, with improving trends in balance since these breaks.
+The authors largely attribute this to equalization of population centers (market sizes), television, and increases in utilization of the international talent pool.
 
 Baseball is not the only sport where this type of analysis has been applied.
-\cite{PalaciosHuerta2004} looked for structural changes in soccer using data from British soccer leagues through 1996.
-They had data from division I (Premier League) and II starting in the late 1800s, and data from lower divisions III and IV from just after WWII starting in 1947.
-They identify a number of change points.
-Notably they identify a change point in the mean of margin of victory in 1925 related to the change in the definition of offsides (changed from 3 players to 2 players), an change points in the variability of number of goals in the early 1980s and 1992 corresponding to the change in number of points for a win (i.e.
-went from 2 points to 3 points) and a change in the backpass rule, respectively.
-\cite{FortLee2007} looked for structural breaks in major North American sports other than baseball (i.e.
-basketball (NBA), hockey (NHL), and American football (NFL)).
+\cite{PalaciosHuerta2004} estimated structural changes in soccer using data from British soccer leagues through 1996.
+Notably they identify a change point in the mean of margin of victory in 1925 related to the change in the definition of offsides (changed from 3 players to 2 players), an change points in the variability of number of goals in the early 1980s and 1992 corresponding both with a change in number of points for a win (from 2 points to 3 points) and a change in the backpass rule.
+\cite{FortLee2007} looked for structural breaks in three other major North American sports leagues: National Basketball Association (NBA), National Hockey League (NHL), and National Football League (NFL).
 They identified a number of change points related to competitive balance in each sport that often, though not always, correspond to league expansion, league mergers, or other major events in a sport (e.g.
@@
 % # WHAT DID THEY FIND?   ADD MORE HERE.  
+Subsequent work has analyzed attendance shifts and trend changes in all four major North American leagues, seeking to understand relationships between league policies – such as free agency and expansion – on both balance and subsequent fan demand (\cite{LeeFort2008,MillsFort2014,MillsFort2018}).
+Many of the largest breaks in demand tended to be coincident with major wars, the Great Depression, expansion, and work stoppages.
+Further empirical analysis has addressed competitive balance in college football and college basketball, finding changes associated with realignment and splits of National Collegiate Athletics Association (NCAA) divisions (\cite{SalagaFort2017, MillsSalaga2015}).
 
@@
 However, recent methodolgical developments in change point analysis allow for change point analysis in {\bf multivariate} time series, which is the focus of this current manuscript.
-This work uses the Double CUSUM Binary Segmentation algorithm (\cite{Cho2016,ChoFryzlewicz2014}) to look for change points in Major League Baseball at two main levels.
-First, we look for structural changes in the league as a whole across teams to empirically define different eras in baseball.
-Second, we look for change points within a team to determine eras of a team.
-This second analysis can be used to identify the beginning and end of so called "dynasties", periods of sustained excellent performance by a team.
-In addition, we can also identify the opposite, sustained periods of poor performance.
+This work uses the Double CUSUM Binary Segmentation algorithm (\cite{Cho2016}) and the Sparsified Binary Segmentation algorithm (\cite{ChoFryzlewicz2014}) to look for change points in Major League Baseball at two main levels.
+First, we look seek to identify structural changes in the league as a whole across teams to empirically define different eras in baseball, providing data-driven empirical context for historical analysis of the sport.
+Second, we look for change points within individual teams to determine relevant eras of team performance and management.
+This second analysis can be used to identify the beginning and end of so called "dynasties", periods of sustained excellent performance by a team, sustained periods of poor performance, and changes to approaches to performance output across eras.
 
@@
 
-\section{Data}
-%EXPLAIN DATA NEEDS FOR 1) ERAS and 2) DYNASTIES.
-We obtained multi-year baseball statistics from the Lahman database (\cite{Friendly2021Lahman}).
-Over the course of the sport's history, numerous statistics have been collected, and for our analysis, we focused on year-end statistics pertaining to the teams.
-In order to implement the DC algorithm, each time series is required to be the same length.
-As a result, when studying eras in our analysis we focus on a set of 16 teams that have all existed over the period from 1900 to 2020 because this period of time covers most of the history of organized professional baseball and there are enough teams that have all existed over this time period to perform meaningful analysis.
-If we had instead chosen to extend this period to, for instance, 1890 to 2020, we would only have had XX teams available.
-This timeframe of $T=120$ years encompassed a total of 16 franchise teams listed in table \ref{tab:120year:franch}.
-The statistics of interest for these teams include runs, hits, home runs, balls, strikeouts, at-bats, stolen bases, number of games played in a season, attendance, runs against, hits against, home runs against, balls against, and strikeouts against.
-However, a few statistics, such as balls, stolen bases, strikeouts, and attendance, had missing data for certain years.
-To address this, we employed multiple imputations using classification and regression trees (\cite{Vanbuuren1999flexible}) to fill in the missing values and ensure consistent data across all analyses.
-(HOW DID WE DO THRE IMPUTATION?
-Was is single imputation?
-)
-
-For the dynasty analysis we used teams over different periods of time for every year that each team has existed....MORE DETAILS.
-
-\begin{table}
-\begin{tabular}{|r|}
-Franchise  \\
-\hline
-NYY Full names 
- NYY, BOS, LAD, ATL, CHW, CHC, CIN, CLE, DET, BAL, SFG, OAK, PHI, PIT, STL, and MIN\\
-\hline
-\end{tabular}
-\caption{}
-\label{tab:120year:franch}
-\end{table}
-
 \section{Methods}
@@
 
-R Pacakge: \cite{hdbinseg}
-
-A commonly employed method for detecting changes over time involves the CUSUM statistic in binary segmentation (CITE XX), which creates a tree diagram, dividing the time series after successfully identifying a change point until no further change points are detected.
+%R Pacakge: \cite{hdbinseg}
+A commonly employed method for detecting changes over time involves the CUSUM statistic in binary segmentation, which creates a tree diagram, dividing the time series after successfully identifying a change point until no further change points are detected.
 The magnitude of the CUSUM difference between two segments generally indicates a potential change point, depending on the assumptions and statistical tests applied to the time series.
@@
 
-To achieve this, the DC statistic from \cite{Cho2016} involves ordered CUSUM values for a given time location, favoring larger values as potential change point candidates.
-It incorporates a partial thresholding of the smaller ordered CUSUM values, effectively representing a scaled average of the most prominent values at a given location across all $n$ (WHY WAS "n" in quotes here?) time series.
+To achieve this, Cho's DC statistic involves ordered CUSUM values for a given time location, favoring larger values as potential change point candidates.
+It incorporates a partial thresholding of the smaller ordered CUSUM values, effectively representing a scaled average of the most prominent values at a given location across all "n" time series.
 Maximizing the DC statistic across all time series and potential time locations results in a test statistic for detecting a change point.
@@
 
-To determine the testing criterion $\pi_{n,T}^\phi$, a Generalized Dynamic Factor Model (GDFM) bootstrapping algorithm is employed (CITE XX), accounting for potential correlations within and between high-dimensional time series.
-This methodology enables the detection of second-order change points, utilizing methodology from \cite{ChoFryzlewicz2014} Haar wavelet periodograms and cross-periodograms, as opposed to the traditional CUSUM statistic.
-For further details, refer to \cite{Cho2016} and \cite{ChoFryzlewicz2014}.
-These methods were implemented using the \texttt{hdbinseg} \cite{hdbinseg} in the R programming language (\cite{R2023language}).
+To determine the testing criterion $\pi_{n,T}^\phi$, a Generalized Dynamic Factor Model (GDFM) bootstrapping algorithm is employed, accounting for potential correlations within and between high-dimensional time series.
+This methodology enables the detection of second-order change points, utilizing methodology from\cite{ChoFryzlewwicz2014} Haar wavelet periodograms and cross-periodograms, as opposed to the traditional CUSUM statistic.
+For further details, refer to \cite{Cho2016} and \cite{ChoFryzlewwicz2014}.
 
-Discuss somewhere how we can find mean change points and ALSO variance change points.
+%SEGUE parageraph Using eras and dynasties:::::::::
 
-SEGUE parageraph Using eras and dynasties:::::::::
-
-Since we are interested in finding where in time these different eras of baseball have occurred, we first examine each baseball statistic of interest as their own analysis for all 16 teams.
-Those baseball statistics are home runs, strikeouts, balls, stolen bases, runs, and attendance which were all standardized to the number of games played in a season since we are interested in seeing how the different statistics change from season to season.
-Using the methodology from \cite{ChoFryzlewicz2014} a thresholding parameter was determined from bootstrapping and then used to find where change points in mean and variance existed in the panel of teams data of a given baseball statistic.
-To examine the idea of changes throughout teams and types of statistics (initially), for each season the average baseball statistic was found from all the teams that year and then an average time series was created of each of the types of statistics of interest.
+%Using the methodology from \cite{ChoFryzlewicz2014} a thresholding parameter was determined from bootstrapping and then used to find where change points in mean and variance existed in the panel of teams data of a given baseball statistic.
+To examine the idea of changes throughout teams and types of statistics, for each season the average baseball statistic was found from all the teams that year and then an average time series was created of each of the types of statistics of interest.
 Then the change point methodology was performed on those 6 time series of average statistic types.
 
-Continuing our investigation, we explored the concept of "dynasties" within modern baseball teams, irrespective of their length of a teams existence.
+%Continuing our investigation, we explored the concept of "dynasties" within modern baseball teams, irrespective of their length of a teams existence.
 A dynasty, in this context, refers to a period characterized by significant success or potentially hardships.
@@
 This approach aligns well with our methodology, as it involves examining panels of time series data encompassing all statistics for a singular team.
-The statistics of interest for this analysis include runs, hits, home runs, balls, strikeouts, runs against, hits against, home runs against, balls against, and strikeouts against.
+%The statistics of interest for this analysis include runs, hits, home runs, balls, strikeouts, runs against, hits against, home runs against, balls against, and strikeouts against.
 To ensure comparability, these statistics were standardized using the season average and standard deviation derived from all teams present in each respective season.
@@
 %hdbinseg: Change-Point Analysis of High-Dimensional Time Series via Binary Segmentation
+
+\subsection{Data}
+%EXPLAIN DATA NEEDS FOR 1) ERAS and 2) DYNASTIES.
+We obtained multi-year baseball statistics from the Lahman database (\cite{Friendly2021Lahman}).
+Over the course of the sport's history, numerous statistics have been collected, and for our analysis, we focused on year-end statistics at the team-level.
+To ensure consistency in our methodology, it was necessary for each team to have existed from 1900 until the end of the dataset in 2020.
+This timeframe of 120 years, denoted as $T$, encompassed a total of 16 franchise teams, namely NYY, BOS, LAD, ATL, CHW, CHC, CIN, CLE, DET, BAL, SFG, OAK, PHI, PIT, STL, and MIN.
+The statistics of interest for these teams include runs (R), hits (H), home runs (HR), base-on-balls (walks; BB), strikeouts (K), at-bats (AB), stolen bases (SB), number of games played in a season (G), attendance , runs against (RA), hits against (HA), home runs against (HRA), base-on-balls against (BBA), and strikeouts against (SO).
+However, a few statistics, such as vase-on-balls, stolen bases, strikeouts, and attendance, had missing data for certain years.
+To address this, we employed multiple imputations using classification and regression trees (cite mice/cart) to fill in the missing values and ensure consistent data across all analyses.
+
+\subsection{Eras and Dynasties}
+\subsubsection{Eras}
+Since we are interested in finding where in time these different eras of baseball have occurred, 
+we first examine each baseball statistic of interest as their own analysis for all 16 teams. Those 
+baseball statistics are home runs, strikeouts, balls, stolen bases, runs, and attendance which were  all standardized to the number of games played in a season since we are interested in seeing how the different statistics change from season to season.
+Using the methodology from Cho and Fry (CITE) a thresholding parameter was determined from bootstrapping and then used to find where change points in mean and variance existed in the panel of teams data of a given baseball statistic.
+To examine the idea of changes throughout teams and types of statistics (initially), for each season the average baseball statistic was found from all the teams that year and then an average time series was created of each of the types of statistics of interest.
+Then the change point methodology was performed on those 6 time series of average statistic types.
+
+\subsubsection{Dynasties}
+Continuing our investigation, we used change point analysis to empirically identify "dynasties"  and "collapses" within modern baseball teams, irrespective of their length of a teams existence.
+A dynasty, in this context, refers to a period characterized by significant success.
+Alternatively, long-term sustained hardships are then defined as a collapse.
+Using this method, a dynasty is identified when a team experiences a noticeable upward shift across all their statistical outputs where as a collapse would be identified through downward shifts at a change point.
+To identify dynasties and collapses, we use runs, hits, home runs, base-on-balls, strikeouts, runs against, hits against, home runs against, base-on-balls against, and strikeouts against.
+As noted earlier, these measures were standardized using the season average and standard deviation derived from all teams present in each respective season.
+By standardizing, we capture a team’s offensive and defensive performance relative to peers within a season.
+
+Notably, if a team exhibits a high and positive offensive performance, it can be balanced to zero by large and negative defensive statistics when standardized.
+Therefore, if a team excels in both aspects of the game, it would have numerous time points above zero, whereas underperforming teams would display many below-zero values.
+For each team, we examined their ten statistics, analyzing both mean and variance changes to identify potential change points.
+By scrutinizing these indicators, we aimed to uncover shifts in a team’s performance and assess their significance within the context of dynasties.
+
+KEEP THE NAMES YOU WANT
+\begin{tabular}{c|c}
+\hline
+Franchise Label & (Modern??) Team Name\\
+\hline
+ATL & Boston Beaneaters\\
+\hline
+ATL & Boston Doves\\
+\hline
+ATL & Boston Rustlers\\
+\hline
+ATL & Boston Braves\\
+\hline
+ATL & Boston Bees\\
+\hline
+ATL & Milwaukee Braves\\
+\hline
+ATL & Atlanta Braves\\
+\hline
+BAL & Milwaukee Brewers\\
+\hline
+BAL & St. Louis Browns\\
+\hline
+BAL & Baltimore Orioles\\
+\hline
+BOS & Boston Americans\\
+\hline
+BOS & Boston Red Sox\\
+\hline
+CHC & Chicago Orphans\\
+\hline
+CHC & Chicago Cubs\\
+\hline
+CHW & Chicago White Sox\\
+\hline
+CIN & Cincinnati Reds\\
+\hline
+CIN & Cincinnati Redlegs\\
+\hline
+CLE & Cleveland Blues\\
+\hline
+CLE & Cleveland Bronchos\\
+\hline
+CLE & Cleveland Naps\\
+\hline
+CLE & Cleveland Indians\\
+\hline
+DET & Detroit Tigers\\
+\hline
+LAD & Brooklyn Superbas\\
+\hline
+LAD & Brooklyn Dodgers\\
+\hline
+LAD & Brooklyn Robins\\
+\hline
+LAD & Los Angeles Dodgers\\
+\hline
+MIN & Washington Senators\\
+\hline
+MIN & Minnesota Twins\\
+\hline
+NYY & Baltimore Orioles\\
+\hline
+NYY & New York Highlanders\\
+\hline
+NYY & New York Yankees\\
+\hline
+OAK & Philadelphia Athletics\\
+\hline
+OAK & Kansas City Athletics\\
+\hline
+OAK & Oakland Athletics\\
+\hline
+PHI & Philadelphia Phillies\\
+\hline
+PHI & Philadelphia Blue Jays\\
+\hline
+PIT & Pittsburgh Pirates\\
+\hline
+SFG & New York Giants\\
+\hline
+SFG & San Francisco Giants\\
+\hline
+STL & St. Louis Cardinals\\
+\hline
+\end{tabular}
 
@@
 
-teams <- read_csv("/Users/gregorymatthews/Dropbox/baseball_cpt/data/Teams.csv") %>% 
+teams <- read_csv("../data/Teams.csv") %>% 
   clean_names()
@@
 
-# teams %>% 
-#   filter(franch_id %in% team_names) %>% 
-#   group_by(franch_id) %>% 
-#   summarise(min(year_id)) ## filter after 1901
-
 old_teams_full <- teams %>% 
   filter(franch_id %in% team_names & year_id > 1900) %>% 
-  select(year_id, franch_id, w, l, r, ab, h, hr, bb, so)
+  select(year_id, franch_id,g, r, ab, sb, hr, bb, so, attendance)
 
-teams_wide <- old_teams_full %>% 
+names(old_teams_full)[-c(1:2)]
+## 16 missing values from SO
+apply(old_teams_full, 2,function(x) sum(is.na(x)))
+
+imp <- mice::mice(teams[c("r","h","hr","bb","so", "ab", "sb", "g", "attendance", 
+                          "ra", "ha", "hra", "bba", "soa")], method = "cart")
+teams_full_nomiss <- complete(imp)
+
+full_teams <- bind_cols(teams[c("year_id", "franch_id")], teams_full_nomiss) %>% 
+  filter(franch_id %in% team_names & year_id > 1900) %>% 
+  mutate(hr_per_game = hr/g, 
+         so_per_game = so/g,
+         bb_per_game = bb/g,
+         att_per_game = attendance/g,
+         sb_per_game = sb/g, 
+         r_per_game = r/g) %>% 
+  select(year_id, franch_id, hr_per_game, so_per_game, bb_per_game, att_per_game, sb_per_game, r_per_game)
+
+# cpt <- function(dataset, var){
+#   wide <- dataset %>% 
+#     pivot_wider(id_cols = franch_id, 
+#                 names_from = year_id, 
+#                 values_from = var)
+#   
+#   ## find thresholding value for cpt using bootstrapping
+#   thrs <- dcbs.thr(as.matrix(wide[-1]), interval = c(1, dim(wide[-1])[2]))
+#   ### if I lower the threshold more changepoints
+#   hr_cpt_dcbs <- dcbs.alg(as.matrix(wide[-1]), cp.type = 1, thr = thrs)
+#   cpt_years <- as.numeric(names(wide[-1])[hr_cpt_dcbs$ecp])
+#   return(list(threshold = thrs,
+#               years = cpt_years),
+#               variable = var)
+#   
+# }
+
+changes <- lapply(names(full_teams)[-c(1:2)], 
+                  function(i){
+                    wide <- full_teams %>% 
                       pivot_wider(id_cols = franch_id, 
                                   names_from = year_id, 
-              values_from = hr)
+                                  values_from = i)
 
                     ## find thresholding value for cpt using bootstrapping
-thrs <- dcbs.thr(as.matrix(teams_wide[-1]), interval = c(1, dim(teams_wide[-1])[2]))
+                    thrs <- dcbs.thr(as.matrix(wide[-1]), interval = c(1, dim(wide[-1])[2]))
                     ### if I lower the threshold more changepoints
-hr_cpt_dcbs <- dcbs.alg(as.matrix(teams_wide[-1]), cp.type = 1, thr = thrs-1)
-hr_cpt_dcbs$ecp
-hr_cpt_sbs <- sbs.alg(as.matrix(teams_wide[-1]), cp.type = 1)         
-hr_cpt_sbs$ecp
+                    cpt_dcbs <- dcbs.alg(as.matrix(wide[-1]), cp.type = 1, thr = thrs)
+                    cpt_years <- as.numeric(names(wide[-1])[cpt_dcbs$ecp])
+                    return(tibble(threshold = thrs,
+                                years = cpt_years,
+                                variable = i))
 
-cpt_years <- as.numeric(names(teams_wide[-1])[hr_cpt_dcbs$ecp])
+                  })
 
-ggplot(old_teams_full, aes(year_id, hr, color = franch_id)) +
+changes_var <- lapply(names(full_teams)[-c(1:2)], 
+                  function(i){
+                    wide <- full_teams %>% 
+                      pivot_wider(id_cols = franch_id, 
+                                  names_from = year_id, 
+                                  values_from = i)
+
+                    ## find thresholding value for cpt using bootstrapping
+                    thrs <- dcbs.thr(as.matrix(wide[-1]), interval = c(1, dim(wide[-1])[2]))
+                    ### if I lower the threshold more changepoints
+                    cpt_dcbs <- dcbs.alg(as.matrix(wide[-1]), cp.type = 2, thr = thrs)
+                    cpt_years <- as.numeric(names(wide[-1])[cpt_dcbs$ecp])
+                    return(tibble(threshold = thrs,
+                                  years = cpt_years,
+                                  variable = i))
+
+                  })
+
+time_cpts <- changes %>% 
+  tibble() %>% 
+  unnest(cols = c(.)) %>% 
+  mutate(change = "mean")
+
+time_cpts_var <- changes_var %>% 
+  tibble() %>% 
+  unnest(cols = c(.)) %>% 
+  mutate(change = "variance")
+
+full_teams %>% 
+  pivot_longer(cols = -c(year_id, franch_id), names_to = "variable", values_to = "value") %>% 
+  ggplot(aes(year_id, value, color = franch_id)) +
   geom_line() +
-  geom_vline(xintercept = cpt_years)
+  geom_vline(data = bind_rows(time_cpts, time_cpts_var), aes(xintercept = years, linetype = change)) +
+  facet_wrap(~ variable, scales = "free_y")
 
@@
 
-%
-% # <<>>=
-% # load("./baseball_cpt/data/teams_cpt_time_df.RData")
-% # teams_cpt_time_df
-% # @
-%
+<<echo = FALSE, message = FALSE, warning = FALSE>>=
+### season average overall
+
+szn_ave <- full_teams %>% 
+  group_by(year_id) %>% 
+  summarise(szn_ave_hr = mean(hr_per_game),
+            szn_ave_so = mean(so_per_game),
+            szn_ave_bb = mean(bb_per_game),
+            szn_ave_sb = mean(sb_per_game)) %>% 
+  pivot_longer(cols = -year_id, names_to = "stat", values_to = "value")
+
+wide <- szn_ave %>% 
+  pivot_wider(id_cols = stat, 
+              names_from = year_id, 
+              values_from = value)
+
+## find thresholding value for cpt using bootstrapping
+thrs <- dcbs.thr(as.matrix(wide[-1]), interval = c(1, dim(wide[-1])[2]))
+                    ### if I lower the threshold more changepoints
+cpt_dcbs <- dcbs.alg(as.matrix(wide[-1]), cp.type = 1, thr = thrs)
+cpt_years <- as.numeric(names(wide[-1])[cpt_dcbs$ecp])
+
+#cpt_years
+
+## variance
+## find thresholding value for cpt using bootstrapping
+thrs_var <- dcbs.thr(as.matrix(wide[-1]), interval = c(1, dim(wide[-1])[2]))
+                        ### if I lower the threshold more changepoints
+cpt_dcbs_var <- dcbs.alg(as.matrix(wide[-1]), cp.type = 2, thr = thrs_var)
+cpt_years_var <- as.numeric(names(wide[-1])[cpt_dcbs_var$ecp])
+
+#cpt_years_var
+
+cpt_szn <- bind_rows(tibble(year = cpt_years, cpt_stat = "mean"), 
+                     tibble(year = cpt_years_var, cpt_stat = "var"))
+
+ggplot(szn_ave) +
+  geom_line(aes(year_id, value, color = stat)) +
+  geom_vline(data = cpt_szn, aes(xintercept = year, linetype = cpt_stat))
+
+@
 
